Evolution of lateralized gustation in nematodes

Version 1

Abstract

Abstract

Animals with small nervous systems have a limited number of sensory neurons that must encode information from a changing environment. This problem is particularly exacerbated in nematodes that populate a wide variety of distinct ecological niches but only have a few sensory neurons available to encode multiple modalities. How does sensory diversity prevail within this neuronal constraint? To identify the genetic basis for patterning different nervous systems, we demonstrate that sensory neurons in the Pristionchus pacificus respond to various salt sensory cues in a manner that is partially distinct from that of the distantly related nematode C. elegans. By visualizing neuronal activity patterns, we show that contrary to previous expectations based on its genome sequence, the salt responses of P. pacificus are encoded in a left/right asymmetric manner in the bilateral ASE neuron pair. Our study illustrates patterns of evolutionary stability and change in the gustatory system of nematodes.

Introduction

Nematodes form a vast array of ecological relationships, from specialized parasite-host dependencies to nematode-microbial interactions, each one demanding exquisitely fine-tuned sets of sensory palates that span multiple modalities (1–8). Yet, the number of sensory neurons across diverse nematode species seems to be constrained (9–11). Several free-living and parasitic nematode species examined by serial section electron microscopy have nearly identical numbers of 12-13 pairs of head sensory neurons, known as the amphid neurons (11–15). How does sensory diversity arise within this neuronal constraint? When coupled with well-described neuronal anatomy, this conserved neuron count allows for detailed comparisons at the single-cell resolution and represents an opportunity to interrogate how sensory cues are processed by anatomically similar nervous systems to produce species-specific or developmental stage-dependent behavioral outputs. To identify the genetic basis for patterning different nervous systems and to understand the processes that underlie evolutionary changes in adapting to different environments, several comparative model systems have been developed to promote comparisons to the well-studied nematode Caenorhabditis elegans at the genetic and cellular levels, including the predatory entomophilic nematode, Pristionchus pacificus (11, 16–19).

As expected from their association with insects in the wild, the olfactory preferences of P. pacificus are distinct from those of C. elegans and the human parasite Strongyloides stercoralis (2, 3, 20), but little is known about P. pacificus responses to water-soluble compounds. In C. elegans, the main salt receptor neuron class comprises of a bilateral pair of left and right ASE neurons (ASEL and ASER), which serve to induce an attractive locomotory response toward an increase in salt concentration (21). The gene che-1 (chemosensory defective) encodes a transcription factor that is exclusively expressed in the ASE neurons and is required for their proper differentiation, such that a che-1 mutant results in defective salt attraction (22–25).

One major role of C. elegans CHE-1 is to promote lateral asymmetry in the ASE neurons. The left and right ASE neurons asymmetrically express receptor-type guanylyl cylases (rGCs, encoded by gcy genes) (26). This finding led to the realization that the ASE neurons are lateralized, such that the left and the right ASE neurons differentially respond to distinct salt ions (27–30). This observation led in turn to the identification of a complex gene regulatory network that genetically programs the distinct sensory potentials of the left and right ASE neurons (31), which includes a miRNA, lsy-6, at the top of this gene regulatory network (31–33). However, the lsy-6 miRNA evolved selectively in the Caenorhabditis genus (34) and is absent in P. pacificus. Moreover, P. pacificus does not show an expansion of the ASEL-type and ASER-type rGCs, as C. elegans does (Hong et al., 2019). With these two genomic observations in mind, we had previously proposed that the ASE neurons are unlikely to be lateralized in P. pacificus (11).

However, we now revise this view in light of our work on mapping chemosensory responses in P. pacificus on the level of behavior and neuronal activity. We demonstrate that P. pacificus does in fact show lateralized chemosensory profiles, indicating that P. pacificus must have evolved independent means to establish ASE laterality. We also show that the tastant palate of P. pacificus is distinct from that of C. elegans, and that its dependence on ASE, as well as its terminal selector transcription factor che-1, have also diverged.

Results

P. pacificus and C. elegans differ in their behavioral responses to salts

Previous cross-species comparisons between P. pacificus and C. elegans indicated strong differences in olfaction preferences that reflect their divergent evolutionary histories and ecology (2). To identify the neurons that mediate gustation, we first compared the chemosensory profiles of these two species toward water-soluble ions. In this survey, we found ammonium salts to be the strongest attractants to wildtype P. pacificus adult hermaphrodites (NH4Br, NH4Cl, NH4I), with NH4I significantly more attractive to P. pacificus compared to C. elegans (Fig. 1). Notably, in contrast to C. elegans (35), we find that P. pacificus is only moderately and weakly attracted to NaCl and LiCl, respectively (Fig. 1). Also, P. pacificus is repulsed by acetate salts (NaAc, NH4Ac), which induce attractive responses in C. elegans (36). We conclude that P. pacificus and C. elegans display differences in their salt preferences.

A comparison of chemotaxis responses to water-soluble ions between P. pacificus and C. elegans.

The J4/L4 larvae and young adult hermaphrodites from the two species responded to NH4I, LiCl, and acetates significantly differently. ****P<0.0001, **P<0.01 significant difference found between wildtype P. pacificus and C. elegans by unpaired t-test. Each assay involves a minimum of 10 J4/adult hermaphrodites.

Ppa che-1 shows similarities and difference to Cel che-1 in both expression and function

The C. elegans che-1 mutant (chemotaxis-defective) was originally isolated for its inability to respond to a broad panel of salt tastants (22, 35), including those described above (29, 30, 36). Che-1 was found to encode for a Zn finger transcription factor that is exclusively expressed in the ASE neuron pair (25), which through laser ablations had been found to be the main salt receptor neurons (37). che-1 was found to control the entire differentiation of the ASE neurons, including the expression of putative receptors of the GCY receptor guanylyl cyclase family (23–25, 38).

To assess whether che-1 performs a similar function in salt perception for P. pacificus as for C. elegans, we analyzed the expression and function of the P. pacificus che-1 ortholog. In a previous paper, we reported that the 5’ region of the sole Pristionchus che-1 ortholog directs reporter expression to the Ppa ASE and Ppa ASG neuron classes (11). Re-examination of our provisional cell identifications using a newly-generated che-1p::GFP strain with stronger neurite expression revealed highly elaborated finger-like dendritic endings in the more anterior amphid neuron that could unambiguously be assigned to the AFD neurons (Fig. 2A-C), prompting us to reassign expression of che-1 to ASE and AFD.

P. pacificus che-1 expression in ASE and AFD amphid neurons.

(A-B) The che-1::GFP marker in the che-1::RCaMP reporter strain is expressed in the ASE and AFD amphid neurons. (C) che-1::GFP expression is detectable in the morphologically distinct AFD neurons with ‘finger’-like dendritic endings. (D-E) Immunostaining of CHE-1::ALFA shows two pairs of amphid neuron cell bodies corresponding to the ASE and AFD neurons; dorsal-ventral view (n=114). (F) The loss of che-1 results in loss che-1::RFP expression in the ASE (circle) while retaining AFD expression. Anterior is left and the scale bar in (C) represents 50 µm for all panels.

We confirmed that the che-1 reporter transgene indicates the full expression of the endogenous che-1 locus by tagging the endogenous che-1 locus with an ALFA-tag (39). The che-1::ALFA animals showed staining in 2 pairs of head neurons whose position is consistent with being the ASE and AFD neurons (Fig. 2D-E). By crossing the che-1 reporter transgene into a che-1 mutant background (see below), we also found that che-1 autoregulates its own expression (Fig. 2F), as it does in C. elegans (25).

Next, we examined whether P. pacificus salt responsiveness shows similar che-1 dependence as in C. elegans. We generated two putative null alleles in the P. pacificus che-1 homolog using CRISPR/Cas9 genome engineering, through introduction of small deletions in the first exon of the gene, thereby resulting in frameshift and premature stops (Fig. 3A). Both Ppa che-1 alleles exhibited defects in attraction towards NH4Cl and LiCl compared to wildtype (Fig. 3B). However, unlike in C. elegans, Ppa-che-1 mutants showed no detectable difference in responses to NH4l, NaCl, and NaAc.

Che-1 expressing amphid neurons are required for sensing water-soluble ions in P. pacificus.

(A) The che-1 locus with CRISPR/Cas9-induced mutations in Exon 1. (B) The che-1 mutants show defects in attraction towards NH4Br, NH4Cl, and LiCl. (C) The che-1p::HisCl animals lose attraction towards NH4Br, NH4Cl, and NH4I in a histamine dependent manner. (D) A schematic map of the P. pacificus AM7(ASE) and AM12(AFD) amphid neurons that express the che-1p::RCaMP used in calcium imaging. The ASE axons are the only amphid axons in P. pacificus to cross each other over the dorsal lateral midline contralaterally. **P<0.01, *P<0.05 Dunnett’s post-hoc comparison showing alleles with significant difference to wildtype P. pacificus (PS312).

We considered two different possibilities for the behavioral differences of Ppa che-1 and Cel che-1 mutants. P. pacificus may use sensory neurons other than ASE to sense these cues, or alternatively, Ppa-che-1 may not have the same fundamental impact on ASE function in P. pacificus as it does in C. elegans. To explore these different possibilities, we silenced che-1 expressing neurons by expressing codon-optimized HisCl channel under control of the che-1 promoter. Histamine-treated che-1p::HisCl animals showed complete loss of attraction to NH4Br, NH4Cl, and NH4I. Histamine treatment did not significantly alter their repulsive response to NaAc (Fig. 3C). These findings corroborate that NaAc is sensed by neurons other than ASE (or AFD, in which the che-1 promoter also drives HisCl). Since NH4I sensation is affected by silencing of che-1(+) neurons but is unaffected in che-1 mutants, ASE differentiation may be more greatly impacted by the silencing of ASE than by the loss of che-1.

The ASE neurons show left/right asymmetric responses to salt

To assess whether Ppa ASE neurons show the same lateralized response to salt ions as Cel ASE neurons, we generated transgenic P. pacificus lines that express RCaMP in ASE neurons and assessed calcium responses to attractive salts (Fig. 4, S1-S2). Specifically, we looked for changes in calcium levels immediately after the addition and removal of specific salts. When 250 mM NH4Cl is administered, an ‘ON’ response is observed as calcium transiently increases in the left ASE neuron (ASEL). In contrast, an ‘OFF’ response was observed as calcium sharply dips in the right ASE neuron (ASER) before quickly returning to baseline when the salt was removed, which suggests hyperpolarization of this neuron (Fig. 4A-B)(29). However, when presented with a 10-fold lower concentration of 25 mM NH4Cl, the ‘OFF’ response completely disappeared in ASER while the ‘ON’ response became more pronounced in ASEL. The ASER responses to 250 mM and 25 mM NaCl (Fig. 4D) were very similar to the ‘OFF’ response (including hyperpolarization) observed for NH4Cl, but the ASEL responses differ between the two salts: instead of the ‘ON’ response expected of ASEL, we observed a relatively weak ‘OFF’ response without the characteristic hyperpolarization but accompanied by an attenuated ‘bump’, a profile that we classify as an ‘OFF-2’ response (Fig. 4C). Finally, we examined the response to NH4l and found that it also elicited laterally asymmetric responses, but yet again distinct from the responses to both NH4Cl and NaCl, with ASEL showing a strong ‘ON’ response, and ASER showing an ‘ON-OFF’ biphasic response to 250 mM NH4l (Fig. 4E-F)(40, 41). Interestingly, the ASER exhibited an ‘OFF’-only response to 25 mM NH4l, which was not observed for the same concentration of NH4Cl and NaCl, and thus may reflect the higher response to NH4l than to NH4Cl and NaCl in the behavior assays. Altogether, P. pacificus ASE neurons clearly show left/right asymmetric responses to salt attractants and these asymmetric responses show similarities and differences to the C. elegans ASE taste neurons (see Discussion).

ASEL and ASER responses to different concentrations of NH4Cl, NaCl, and NH4I.

Average percent change in RCaMP fluorescence (dF/F) over time (seconds) of tracked ASE (left and right) sensory neurons in P. pacificus. Salts were presented at 10s (“ON”, left vertical line) for a duration of 20s, and then removed (“OFF”, right vertical line) for the remaining 30s; the total recording time was 60s. (A-B) ASEL and ASER neuron responses to high (250 mM, red) compared to low (25 mM, blue) concentrations of NH4Cl. (C-D) ASEL and ASER neuron responses to NaCl. (E-F) ASEL and ASER neuron responses to NH4I. Shaded ribbons represent 95% confidence intervals. Shaded ribbons in (E-F) have been cropped to maintain consistent y-axes across the plots, allowing for easier comparison. Sample sizes are indicated (n).

The AFD neurons also respond to salts in P. pacificus

The RCaMP line that we used to assess calcium responses in ASE is also expressed in AFD, allowing us to simultaneously examine the calcium responses in the AFD (AM12) neurons (Fig. 5, S1, S3-S5). Surprisingly, we detected a distinctly ‘ON-OFF’ biphasic response to all 3 salt types at both concentrations. Specifically, although we observed weaker or comparable responses in AFD neurons when compared to either ASE neuron’s response towards 250 mM NH4Cl, NaCl and NH4l (Fig. 5A-B, 5E-F, 5I-J), the AFD responses were more robust than ASER toward 25 mM NH4Cl, NaCl, and NH4I (Fig. 5D, 5H, 5L). Specifically, AFD neurons responded strongly to 25 mM NaCl when neither one of the ASE neurons showed a positive response (Fig. 5G-H). Averaging the calcium transients separately by AFD left versus right did not result in significant differences in the shape of the neuronal calcium responses, with the exception of the AFDR responses to higher versus lower concentrations of NH4I (Fig. S4-S5). We have not further pursued whether these AFD responses are a reflection of a direct perception of salt or a secondary consequence of communication of salt-perceptive neurons (like ASE) to AFD.

Combined AFD responses in comparison to ASE left or right neuron responses to NH4Cl, NaCl, and NH4I.

Average percent change in RCaMP fluorescence (dF/F) over time (seconds) as described in Fig. 4. Averaged combined AFD (left and right neurons, orange) compared to left or right ASE (green) responses to (A-B) 250mM NH4Cl (C-D) 25mM NH4Cl (E-F) 250mM NaCl (G-H) 25mM NaCl (I-J) 250mM NH4I (K-L) 25mM NH4I. Shaded ribbons represent 95% confidence intervals. Shaded ribbons in (J-K) have been cropped to maintain consistent y-axes across the plots, allowing for easier comparison. Sample sizes are indicated (n).

A target of che-1, the guanylyl cyclase gcy-22.3 is required for chemotaxis

We further explored the asymmetric salt perception by the ASE neurons, which in C. elegans is largely mediated through distinct receptor-type guanylyl cyclases (rGC proteins, encoded by gcy genes), which confer salt specificity via their extracellular domains (Ortiz 2009, Smith 2013). Our previous genome survey of Ppa homologs of gcy genes has revealed patterns that made us question whether Ppa gcy genes are convincing candidates for lateralized chemotactic responses. Specifically, we noted that ASER-expressed C. elegans gcy genes and ASEL-expressed C. elegans gcy genes have only expanded in the Caenorhabditis genus (11). One outlier to this pattern is the Cel gcy-22 gene, which is expressed in ASER, but has not expanded in C. elegans (11, 42). However, this gene has duplicated several times in P. pacificus, resulting in 5 putative Ppa-gcy-22 paralogs (11). We fused the promoter of one of these paralogs, Ppa-gcy-22.3, to gfp and found that transgenic animals express GFP exclusively in ASER (Fig. 6A), identical to the C. elegans gcy-22 ortholog (42). We confirmed its expression in ASER by analyzing animals that carry both the Ppa gcy-22p::gfp reporter and the Ppa che-1p::rfp reporter, showing a unilateral overlap of these reporters in ASER (Fig. 6B).

The laterally asymmetric expression of gcy-22.3 is dependent on the zinc finger transcription factor CHE-1.

(A) The gcy-22.3::GFP marker is expressed exclusively in the right ASE neuron (ASER)(n>200). (B) The gcy-22.3::GFP marker co-localizes with che-1::RFP expression in the ASER. (C) gcy-22.3::GFP expression is absent in the che-1(ot5012) mutant (n=55). Anterior is left and dorsal is top. Scale bar in (A) represents 50 µm for all panels.

To assess whether Ppa-gcy-22.3 is a potential effector of Ppa che-1 function, we crossed the gcy-22.3 reporter into che-1(ot5012) mutants. We found expression of gcy-22.3 was eliminated (Fig. 6C), leading us to conclude that gcy-22.3 is a potential effector of che-1 function, identical to its homolog in C. elegans.

To determine whether and which of the observed salt responses is mediated by the ASER-expressing gcy-22.3, we generated a putative gcy-22.3 null mutant through CRISPR/Cas9 genome editing (2 bp complex deletion that introduces frameshift)(Fig. 7A) and examined its response to the higher salt concentration. The ‘OFF’ response to 250 mM NH4Cl was notably abolished in the ASER neuron in the loss-of-function gcy-22.3 mutant, while the ‘ON’ response in the ASEL remained intact (Fig. 7B-C). However, the responses to 250 mM NaCl were not significantly reduced in the either the ASEL or ASER neuron in the gcy-22.3 mutant (Fig. 7D-E). Furthermore, the gcy-22.3 mutation also reduced the ‘ON’ portion of the AFD biphasic response following presentation of 250 mM NH4Cl and NaCl (Fig. 7F-G). Our findings suggest GCY-22.3 is possibly a P. pacificus salt receptor that can affect the physiological responses to salt in the ASER gustatory neuron, as well as other gustatory neurons that do not express this receptor (AFDs).

ASE and AFD responses in wildtype compared to gcy-22.3 mutants.

Average percent change in RCaMP fluorescence (dF/F) over time (seconds) as described in Fig. 4. (A) WormBase JBrowse view of the gcy-22.3 locus with CRISPR/Cas9-induced mutations in Exon 5. (B-C) ASEL and ASER neuron responses to 250 mM NH4Cl wildtype (gray) and gcy-22.3 mutants (magenta). (D-E) ASEL and ASER neuron responses to 250 mM NaCl in wildtype (gray) and gcy-22.3 mutants (magenta). AFD (combined left and right neurons) responses (F) to 250 mM NH4Cl and (G) to 250 mM NaCl in wildtype (orange) and gcy-22.3 mutants (magenta). Shaded ribbons represent 95% confidence intervals. Sample sizes are indicated (n). For ASEL/R comparisons, bar plots represent the difference between the minimum % dF/F value 10s pre-stimulus and maximum % dF/F value 10s post-stimulus for either the (B) ON or (C-E) OFF stimulus. For AFD comparisons, bar plots represent maximum % dF/F values 10s after the ON and OFF stimulus. *P<0.05, **P<0.01, ***P<0.001 indicate significant difference. “ns” indicate no significant difference. Comparisons between different genotypes were analyzed using unpaired t-test or Mann-Whitney test; comparisons within the same genotype were analyzed using paired t-test.

Discussion

Our study has revealed several insights into the substrates of evolutionary changes between two distantly related nematode species, P. pacificus and C. elegans. We used intracellular calcium levels as our readout for neuronal activity with a genetically encoded calcium sensor in two pairs of che-1-expressing amphid sensory neurons– the first calcium imaging study in P. pacificus. We show that three neuron types (ASE left, ASE right and AFD neurons) each have distinct calcium responses to specific ion concentrations, revealing diversity at the single neuron level. We have identified the first laterally asymmetric marker in P. pacificus, gcy-22.3p::GFP, with its expression limited to ASER homolog (AM7). Our unexpected discovery that neuronal asymmetry is present in the ASE homologs between two distantly related nematode species, despite the lack of a lsy-6 homolog, suggests that functional lateralization in P. pacificus may be mediated by a different genetic pathway compared to C. elegans.

P. pacificus and C. elegans have diverged taste palates

We have shown that while C. elegans is attracted to acetate salts, P. pacificus avoids these acetates. Previous studies have shown that ammonium acetate (NH4Ac) is sensed both as a water-soluble compound as well as a volatile odorant and is mediated by different signaling pathways (36). Like C. elegans, it is therefore likely that ammonium and acetate ions involve a different set of neurons in P. pacificus (i.e. non-che-1 expressing neurons), based on the finding that histamine-treated che-1p::HisCl animals did not significantly attenuate their repulsive response to NaAc. We find that sodium chloride is a common attractant for both P. pacificus and C. elegans, although the magnitude of response is lower than previously published results (29). This result is likely due to differences in generating the salt gradients. Nevertheless, we find that P. pacificus neurons have a distinct response to this salt when compared to those observed in C. elegans neurons. The P. pacificus ASEL neuron responds to a decrease in NaCl concentration (likely sodium), as evidenced by the ‘OFF-2’ response profile, whereas the C. elegans ASEL neuron responds to an increase in sodium concentration. However, in both nematodes, the ASER neuron responds to a decrease in chloride concentration. Additionally, the P. pacificus ASER neuron exhibits a unique ON-OFF response to ammonium iodide, whereas in C. elegans, no ON-OFF type of response is seen in the ASE neurons– the C. elegans ASER neuron responds only to a decrease in iodide concentration. Collectively, our findings highlight the divergence of the P. pacificus salt-sensory neurons compared to those observed in C. elegans.

P. pacificus ASE neurons have narrower sensitivity range

The sensitivity range of the P. pacificus ASER responses is significantly less compared to the ASEL responses, as well as to C. elegans ASER responses. Whereas the C. elegans ASER has a 40-fold sensitivity range in the ‘OFF’ response to the removal of various concentrations of NaCl (1-40 mM)(29, 43–45), the P. pacificus ASER showed the ‘OFF’ response only to 250 mM NH4Cl but not to a 10-fold reduction concentration of NH4Cl (25 mM). For the ASEL in contrast, the response to 25 mM was just as strong as to 250 mM NH4Cl (10-fold) and comparable to the 8-fold concentration range observed for C. elegans ASEL towards NaCl. Alternatively, the magnitude of these sensitivity differences may be also partially due to differences among calcium indicators (i.e. GCaMP and Cameleon), but multiple P. pacificus che-1p::GCaMP strains did not exhibit sufficient basal fluorescence to allow for image tracking and direct comparison. The narrower sensitivity range in P. pacificus chemosensation was also observed for attraction to volatile odors (2), which span only 10-fold, versus up to 10,000-fold in attractive odors for C. elegans (21).

P. pacificus taste neurons exhibit a unique biphasic response

Although the left-ON and right-OFF responses are conserved in the ASE neurons in both species, the biphasic response of ASER to 250 mM NH4I has not been observed in either of the ASE neurons towards salts. In C. elegans, hyperosmotic stimulus such as 1M glycerol, or high concentrations or long duration of CuSO4 exposure both result in biphasic responses by the ASH neurons that sense noxious chemicals (41, 46). Olfactory neurons that mediate avoidance behavior such as the AWB neurons can also exhibit a biphasic response to the presence and removal of a high concentration isoamyl alcohol that normally elicits avoidance behavior (47). The neuron-specific response can also be dependent on the concentration of the chemical compound, since the ASER to 25 mM NH4I was a weak ‘OFF’ rather than a biphasic one. In contrast to the salt-dependent response types by P. pacificus ASE neurons, the P. pacificus AFD neurons also show exclusively biphasic responses with various amplitudes. Further characterization will help determine if biphasic responses can also be found in other sensory neuron types in P. pacificus, specifically those neurons mediating avoidance behavior.

AFD are potentially polymodal neurons

Broadly speaking, C. elegans chemosensory neurons have been classically characterized as specialized neurons for dedicated modalities such as water-soluble chemicals (ASE), volatile odorants (AWA, AWB, AWC), noxious chemicals (ASH), pheromones (ADL, ADF, ASK), as well as light (ASJ) and temperature (AFD)(1, 37, 48– 51). Advances in multi-neuron calcium recordings have since shown that a given odor within a certain concentration range is detected by different ensembles of the 12 amphid neuron classes, including the AFD neurons (52–54). Unexpectedly, we found that the P. pacificus AFD neurons exhibit a distinctive biphasic response to all three salts tested (NH4Cl, NaCl, and NH4I), which differ from the C. elegans AFD calcium responses (54, 55). Moreover, the loss of the receptor gcy-22.3 reduced the AFD ‘ON’ response to NH4Cl and NaCl, indicating ASER has only a limited contribution to the AFD response. The strong positive AFD response to 25 mM NaCl that is absent in both ASE neurons further supports the likelihood that AFD receives inputs, possibly neuropeptidergic, from other amphid neurons. The integration of thermosensation and chemosensation is important for memory-regulated behavior. In C. elegans, maximum chemotaxis indices toward NH4Cl occurs when there is concordance between cultivation temperature and assay temperature (56). The C. elegans AFD neurons are also important for gustatory aversive learning in NaCl avoidance (45). Given the influence of environmental temperature on the P. pacificus mouth-form plasticity and the wide range of micro-climates that wild strains P. pacificus have been isolated from (57–59), temperature and taste preferences could be regulated at multiple genetic levels during crucial developmental decisions.

Changes in the gene regulatory architecture of sensory neuron specification

We found that the key regulator of Cel ASE identity, che-1, is also expressed in Ppa ASE and may play a similar role as a terminal selector in this neuron, based on its effect on ASE-mediated behavior and regulation of the Ppa-gcy-22.3 gene. However, the stronger behavioral effect of silencing of che-1 expressing neurons compared to a che-1 mutant background could either indicate that che-1 does not have as broad a role in controlling ASE differentiation in P. pacificus versus C. elegans. It is also possible that a developmental loss of ASE differentiation may result in compensatory changes in the chemosensory system during early development, as was observed in the C. elegans mating pheromone response by males (60).

Unexpectedly, we found that unlike in C. elegans, the Ppa che-1 gene is also expressed in the AFD neurons. Since we cannot record neural activity in AFD in a che-1 mutant (the che-1p::RCaMP driver fails to be expressed properly in che-1 mutants due to autoregulation), and do not yet have molecular markers for Ppa AFD neurons, we cannot assess whether che-1 affects AFD neuron differentiation.

Perhaps the most striking difference in the gene regulatory architecture of ASE neuron specification is the apparent lack of the key regulator of ASE asymmetry in P. pacificus, the miRNA lsy-6. In C. elegans, the expression of lsy-6 exclusively in ASEL is prepatterned via an early embryonic Notch signal (33) and serves to downregulate the homeodomain transcription factor cog-1 in the ASEL neuron (61). Through a network of downstream regulatory events, asymmetry of rGCs eventually becomes established (31). cog-1 and several asymmetrically expressed downstream effectors of cog-1, such as the die-1, lim-6 and fozi-1 transcription factors are conserved in P. pacificus, but whether the function of these factors in controlling P. pacificus ASE laterality is conserved remains to be determined. In this context it is intriguing to note that another prominent gene regulatory pathway that is controlled by miRNAs in C. elegans, the heterochronic pathway (let-7s, lin-4), appears to have diverged in P. pacificus as well, despite the conservation of the overall physiological readouts of this pathway (temporal patterning of cell lineage divisions)(62). It is tempting to speculate that miRNA-meditated regulatory process is particularly labile.

In conclusion, our work illustrates how comparative behavioral and genetic analyses in nematodes is a powerful strategy to uncover substrates of evolutionary change in simple nervous systems.

Materials and Methods

Nematode Strains

P. pacificus and other nematode strains were maintained at ∼20°C on NGM plates seeded with E. coli OP50 for food as described previously (63); these are derived from standard C. elegans culture methods (64). P. pacificus and other nematode strains used are listed in Table S1.

Chemotaxis assays

The assay for assessing response to salt gradients was adapted from C. elegans and P. pacificus chemotaxis assays (2, 21, 42). Overnight salt gradients were established on 10 cm chemotaxis plates containing 20 ml agar (5 mM KPO4, 1 M CaCl2, 3% Bacto-agar, 1 mM MgSO4) by adding 10 µl of 2.5 M salt solutions for 16 hours. Then, another 4 µl of the same salt solution or water control was added to reinforce the gradient 4 hours before the assay. Just prior to the assay, 1.5 µl of 1 M sodium azide was added to both the attractive salt (A) and the control (C) spots. Young P. pacificus adult hermaphrodite from near-saturated cultures were washed 3x with distilled water and collected by centrifuging at 2000 rpm for 2 minutes. Approximately 200 worms were loaded onto the edge of each assay plate between the gradient sources, and at least 10 combined worms have to reach the scoring arenas to be considered. At least 12 assays constituted each experimental trial, and multiple trials were conducted and averaged for each condition. The Chemotaxis Index (CI) for each end-point assay plate is defined as (A - C)/(A + C). To conduct conditional knockdowns of neurons, 5 M histamine dihydrochloride (Sigma Aldrich H7250) stock solution in sterilized deionized water (Arrowhead CA) was filter-sterilized and top plated onto the agar plates to a final histamine concentration of 25 mM histamine approximately 10 minutes before loading the worms to commence the assay. Most assays lasted ∼3.5 hours at room temperature in line with the speed of P. pacificus without food, with ∼40% of the animals participating. When using the Ppa-che-1p::optHisCl(csuEx83) strain, only animals expressing Ppa-egl-20p::RFP tail marker from the extrachromosomal array were scored. Because of P. pacificus’ strong aversion to acetate, we could not easily assess the individual contributions of salt ions in a saturated background of ammonium acetate as conventionally practiced in C. elegans studies (30).

Ppa-che-1p::HisCl strain

To make the Ppa-che-1p::optHisCl, the P. pacificus codon-optimized histamine-gated chloride channel sequence used in C. elegans (HisCl)(65, 66) was designed using (https://hallemlab.shinyapps.io/ Wild_Worm_Codon_Adapter/) and custom synthesized (Twist Bioscience), and subsequently inserted behind the che-1 promoter (3.1 kb containing the first exon and intron)(11) to create the pHC30 plasmid construct. This Ppa-che-1p::optHisCl plasmid (2 ng/µl) along with PS312 genomic DNA (80 ng/µl) and Ppa-egl-20p::RFP co-injection marker (2 ng/µl) were individually digested with HindIII and assembled as the injection mix to create csuEx83.

Ppa-che-1p::RCaMP reporter strain

To make the Ppa-che-1p::optRCaMP, we generated a transgenic worm strain expressing the codon-optimized genetically-encoded calcium indicator (GECI), jRCaMP1a, in the neurons of interest (67, 68). jRCaMP1a is an improved red GECI based on mRuby with comparable sensitivity to GCaMP6 (69). The codon-optimized RCaMP sequence was custom synthesized (Twist Bioscience), and then subcloned using the plasmid pJET1.2/blunt (Thermo Fisher Scientific) to create the pMM2 plasmid construct. Using Gibson Assembly (E2611, New England Biolab), the codon-optimized RCaMP sequence was introduced downstream of the Ppa-che-1 promoter sequence to create the pMM5 plasmid construct. Separately, we also generated a Ppa-che-1p::optGFP transcriptional reporter with codon-optimized GFP (pMM3) to enable the localization of the che-1-expressing neurons during video acquisition (70). The Ppa-che-1p::optRCaMP (2 ng/µl) and Ppa-che-1p::optGFP constructs (1 ng/µl), along with PS312 genomic DNA (80 ng/µl) and Ppa-egl-20p::RFP (1.5 ng/µl) were individually digested with HindIII and assembled as the injection mix to create csuEx93. Despite multiple attempts, we were unable to generate an equivalent che-1p::GCaMP transgenic line with sufficient basal level of GCaMP expression for a comparison to the RCaMP calcium dynamics.

CRISPR mutagenesis generated mutants

CRISPR/Cas9 mutagenesis was used to generate mutations (70, 71). crRNA and primer sequences, and induced mutations, are included in Table S3 and Fig. S6.

che-1 alleles (PPA01143)

Target crRNA, tracrRNA, and Cas9 nuclease were purchased from IDT Technologies (San Diego, CA). crRNA and tracrRNA were hydrated to 100 µM with IDT Duplex Buffer, and equal volumes of each (0.61 µl) were combined and incubated at 95°C for 5 minutes, then 25°C for 5 minutes. Cas9 protein (0.5 µl of 10 µg/µl) was added, then the mix was incubated at 37°C for 10 minutes. Ppa-egl-20p::RFP was used as a co-injection marker. To reach a final total volume of 40 µl, the Cas9-crRNA-tracrRNA complex was combined with pZH009 (Ppa-egl-20p::RFP) DNA to reach 50 ng/µl final concentration using nuclease-free water. F1 progeny were screened for the presence of Ppa-egl-20p::RFP expression in the tail and candidate F1‘s were sequenced to identify heterozygotes (71). ot5012 has a 4 bp insertion while ot5013 has an 8 bp complex insertion/deletion, and both mutations cause frameshift mutations and premature stop codons. Each allele was outcrossed two times to wildtype before characterization.

gcy-22.3 alleles (PPA04454)

Target crRNA, tracrRNA, and Cas9 nuclease were purchased from IDT Technologies (San Diego, CA). crRNA (RHL1400) and tracrRNA were hydrated to 100 µM with IDT Duplex Buffer, and equal volumes of each (0.61 µl) were combined and incubated at 95°C for 5 minutes, then 25°C for 5 minutes. Cas9 protein (0.5 µl of 10 µg/ µl) was added, then the mix was incubated at 37°C for 10 minutes. Ppa-egl-20p::RFP was used as a co-injection marker. To reach a final total volume of 40 µl, the Cas9-crRNA-tracrRNA complex was combined with pZH009 (Ppa-egl-20p::RFP) DNA to reach 50 ng/µl final concentration using nuclease-free water. F1 progeny were screened for the presence of Ppa-egl-20p::RFP expression in the tail and candidate F1‘s were sequenced to identify heterozygotes (71). csu181 has a 2 bp complex deletion, while csu182 has a 22 bp complex deletion, both mutations cause frameshifts and premature stop codons. Each allele was outcrossed two times to wildtype before characterization.

CHE-1:ALFA C-terminal tagging and immunostaining

For induction of site-specific insertions via CRISPR/Cas-9 mediated mutagenesis, target crRNA, tracrRNA, and Cas9 nuclease were treated as described above. Single-stranded DNA repair template containing the ALFA nanobody tag (RHL1551) with 35 bp of homology arms on the 5’ and 3’ sides were purchased from IDT. To minimize sequence identity, the two copies of the ALFA sequence contain silent mutations. The crRNA (RHL1396) and tracrRNA were hydrated to 100 µM with IDT Duplex Buffer, and equal volumes of each (0.61 µl) were combined and incubated at 95°C for 5 minutes, then 25°C for 5 minutes. F1 animals expressing the co-injection marker egl-20::optRFP were lysed and checked by PCR for insertions.

Non-starved healthy cultures of ALFA-tagged CHE-1 (csu226[Ppa-che-1::2xALFA]) (from six 6 cm plate cultures) were washed with M9 and filtered (Sartorius 84g/m2, Grade 392) and processed as previously described (39). In brief, mixed stage worms were fixed overnight at 4°C on a nutator with 500 µl fixation buffer (4% paraformaldehyde in PBS). The worms were then incubated overnight at 37°C with 500 µl 4% β-mercaptoethanol dissolved in 1% Triton-X100 in 0.1M Tris pH 7.4, and digested in 200 μl of collagenase buffer with 200 units of collagenase type IV (Gibco, Thermo Fisher Scientific, Waltham, MA, USA) at 37°C for ∼3.5 hours. The partially digested worms were subsequently washed three times with 500 µl PBST. After collagenase treatment the centrifugation between the washes was done at low speed, 1000 RCF. After washes, the worms were stained in 50 µl 1% BSA in PBST with the primary antibody (1:100 FluoTag-X2 anti-ALFA-AZdye568, N1502, NanoTag Biotechnologies, Göttigen, Germany), overnight at 4°C in a nutator. Following the washes, worms were resuspended in 50 µl VectaShield mounting medium (H-1000, Vector Laboratories, USA) containing DAPI (Molecular Probes, Thermo Fisher Scientific) and gently mounted on freshly prepared 3% Noble agar pads. Images were acquired on a Leica DM6000 microscope.

Calcium Imaging

To conduct calcium imaging, worms were trapped and imaged within a microfluidic PDMS chip as previously described while delivering stimuli directly to the nose of an immobilized worm (46, 72). The microfluidic chip was connected to a programmable valve controller (ValveBank) that enables the user to toggle between “stimulant ON” and “stimulant OFF” states. Specifically, the ValveBank allows controlled switching of flow from the two outer buffer channels in the chip such that either the control solution (stimulant OFF) or the stimulant solution (stimulant ON) flows over the worm nose (Fig. S7). Buffer solution (for the outer two channels) consisted of M9 buffer with 0.1% Tween-20 and 1 µg/ml fluorescein (73, 74). The worm loading solution consisted of M9 buffer with 0.1% Tween-20 and 1.5 mM tetramisole hydrochloride to immobilize the animal. Water-soluble stimulant solutions consisted of the following water-soluble compounds dissolved in nanopure/milli-Q water: ammonium chloride (NH4Cl), ammonium iodide (NH4I), and sodium chloride (NaCl) at concentrations of 250 mM and 25 mM. 750 mM NaCl elicited very inconsistent responses. The water-soluble stimulant solution was made by diluting the water-soluble compound in nanopure/milli-Q water with 0.1% Tween-20. The control solution consisted of 0.1% Tween-20 nanopure/milli-Q water. Worms were exposed to the control solution, stimulant solution, and then control solution using a 60-second program: 10 seconds stimulant OFF, 20 seconds stimulant ON, and 30 seconds stimulant OFF. As a negative control, animals were exposed only to the control solution, without switching channels, for the duration of the recording. Images were captured using a Zeiss Axio Observer Z1 inverted fluorescence microscope and a pco.panda 4.2 SCMOS camera. Changes in fluorescence intensity were measured in the neurons of interest while the worm was exposed to green light. Images were processed using MetaMorph software version 7.10.5.476. Images were captured at 500 ms exposure time at 2 fps because the baseline fluorescence in Ppa-che-1p::optRCaMP worms was too dim to capture viable data using the standard 100 ms exposure time. Baseline F₀ was measured as the average background-subtracted fluorescence from the first 9 seconds of each recording and change in fluorescence intensity was calculated as dF/F = (F - F₀) / F₀, as described (74). Data were analyzed and plotted using custom scripts generated in MATLAB versions R2021a–R2024a. This code is available at https://github.com/honglabcsun/Calcium-Imaging.

For bar plot comparisons between wildtype and gcy-22.3 mutants, we calculated minimum pre-stimulus and maximum post-stimulus % dF/F values using custom scripts generated in MATLAB. Min-Max data were exported as text files, manually converted and organized into an XLSX file (Microsoft Excel Office16) and imported into Prism GraphPad software (version 10) for data visualization and statistical analysis.

Nomenclature

Throughout the results section, P. pacificus genes will be referred to without the Ppa- prefix; if necessary for comparison to another species such as C. elegans (Cel-) will the Ppa- prefix be used.

eLife Assessment

Mackie and colleagues present a valuable comparison of lateralized gustation in two well-studied nematodes. The evidence they present that ASEL/R lateralization exists and is achieved by different means in P. pacificus compared to C. elegans is solid, but would be strengthened by the inclusion of calcium imaging data for C. elegans and additional verification of P. pacificus AFD neuron identity. This work will be of interest to evolutionary biologists, ecologists, and sensory neurobiologists interested in how small nervous systems make sense of the environment, and how evolution can take multiple paths to asymmetry within a neuron class.

Reviewer #1 (Public review):

Summary:

Mackie and colleagues compare chemosensory preferences between C. elegans and P. pacificus, and the cellular and molecular mechanisms underlying them. The nematodes have overlapping and distinct preferences for different salts. Although P. pacificus lacks the lsy-6 miRNA important for establishing asymmetry of the left/right ASE salt-sensing neurons in C. elegans, the authors find that P. pacificus ASE homologs achieve molecular (receptor expression) and functional (calcium response) asymmetry by alternative means. This work contributes an important comparison of how these two nematodes sense salts and highlights that evolution can find different ways to establish asymmetry in small nervous systems to optimize the processing of chemosensory cues in the environment.

Strengths:

The authors use clear and established methods to record the response of neurons to chemosensory cues. They were able to show clearly that ASEL/R are functionally asymmetric in P. pacificus, and combined with genetic perturbation establish a role for che-1-dependent gcy-22.3 in in the asymmetric response to NH4Cl.

Weaknesses:

The mechanism of lsy-6-independent establishment of ASEL/R asymmetry in P. pacificus remains uncharacterized.

Reviewer #2 (Public review):

Summary:

In this manuscript, Mackie et al. investigate gustatory behavior and the neural basis of gustation in the predatory nematode Pristionchus pacificus. First, they show that the behavioral preferences of P. pacificus for gustatory cues differ from those reported for C. elegans. Next, they investigate the molecular mechanisms of salt sensing in P. pacificus. They show that although the C. elegans transcription factor gene che-1 is expressed specifically in the ASE neurons, the P. pacificus che-1 gene is expressed in the Ppa-ASE and Ppa-AFD neurons. Moreover, che-1 plays a less critical role in salt chemotaxis in P. pacificus than C. elegans. Chemogenetic silencing of Ppa-ASE and Ppa-AFD neurons results in more severe chemotaxis defects. The authors then use calcium imaging to show that both Ppa-ASE and Ppa-AFD neurons respond to salt stimuli. Calcium imaging experiments also reveal that the left and right Ppa-ASE neurons respond differently to salts, despite the fact that P. pacificus lacks lsy-6, a microRNA that is important for ASE left/right asymmetry in C. elegans. Finally, the authors show that the receptor guanylate cyclase gene Ppa-gcy-23.3 is expressed in the right Ppa-ASE neuron (Ppa-ASER) but not the left Ppa-ASE neuron (Ppa-ASEL) and is required for some of the gustatory responses of Ppa-ASER, further confirming that the Ppa-ASE neurons are asymmetric and suggesting that Ppa-GCY-23.3 is a gustatory receptor. Overall, this work provides insight into the evolution of gustation across nematode species. It illustrates how sensory neuron response properties and molecular mechanisms of cell fate determination can evolve to mediate species-specific behaviors. However, the paper would be greatly strengthened by a direct comparison of calcium responses to gustatory cues in C. elegans and P. pacificus, since the comparison currently relies entirely on published data for C. elegans, where the imaging parameters likely differ. In addition, the conclusions regarding Ppa-AFD neuron function would benefit from additional confirmation of AFD neuron identity. Finally, how prior salt exposure influences gustatory behavior and neural activity in P. pacificus is not discussed.

Strengths:

(1) This study provides exciting new insights into how gustatory behaviors and mechanisms differ in nematode species with different lifestyles and ecological niches. The results from salt chemotaxis experiments suggest that P. pacificus shows distinct gustatory preferences from C. elegans. Calcium imaging from Ppa-ASE neurons suggests that the response properties of the ASE neurons differ between the two species. In addition, an analysis of the expression and function of the transcription factor Ppa-che-1 reveals that mechanisms of ASE cell fate determination differ in C. elegans and P. pacificus, although the ASE neurons play a critical role in salt sensing in both species. Thus, the authors identify several differences in gustatory system development and function across nematode species.

(2) This is the first calcium imaging study of P. pacificus, and it offers some of the first insights into the evolution of gustatory neuron function across nematode species.

(3) This study addresses the mechanisms that lead to left/right asymmetry in nematodes. It reveals that the ASER and ASEL neurons differ in their response properties, but this asymmetry is achieved by molecular mechanisms that are at least partly distinct from those that operate in C. elegans. Notably, ASEL/R asymmetry in P. pacificus is achieved despite the lack of a P. pacificus lsy-6 homolog.

Weaknesses:

(1) The authors observe only weak attraction of C. elegans to NaCl. These results raise the question of whether the weak attraction observed is the result of the prior salt environment experienced by the worms. More generally, this study does not address how prior exposure to gustatory cues shapes gustatory responses in P. pacificus. Is salt sensing in P. pacificus subject to the same type of experience-dependent modulation as salt sensing in C. elegans?

(2) A key finding of this paper is that the Ppa-CHE-1 transcription factor is expressed in the Ppa-AFD neurons as well as the Ppa-ASE neurons, despite the fact that Ce-CHE-1 is expressed specifically in Ce-ASE. However, additional verification of Ppa-AFD neuron identity is required. Based on the image shown in the manuscript, it is difficult to unequivocally identify the second pair of CHE-1-positive head neurons as the Ppa-AFD neurons. Ppa-AFD neuron identity could be verified by confocal imaging of the CHE-1-positive neurons, co-expression of Ppa-che-1p::GFP with a likely AFD reporter, thermotaxis assays with Ppa-che-1 mutants, and/or calcium imaging from the putative Ppa-AFD neurons.

(3) Loss of Ppa-che-1 causes a less severe phenotype than loss of Ce-che-1. However, the loss of Ppa-che-1::RFP expression in ASE but not AFD raises the question of whether there might be additional start sites in the Ppa-che-1 gene downstream of the mutation sites. It would be helpful to know whether there are multiple isoforms of Ppa-che-1, and if so, whether the exon with the introduced frameshift is present in all isoforms and results in complete loss of Ppa-CHE-1 protein.

(4) The authors show that silencing Ppa-ASE has a dramatic effect on salt chemotaxis behavior. However, these data lack control with histamine-treated wild-type animals, with the result that the phenotype of Ppa-ASE-silenced animals could result from exposure to histamine dihydrochloride. This is an especially important control in the context of salt sensing, where histamine dihydrochloride could alter behavioral responses to other salts.

(5) The calcium imaging data in the paper suggest that the Ppa-ASE and Ce-ASE neurons respond differently to salt solutions. However, to make this point, a direct comparison of calcium responses in C. elegans and P. pacificus using the same calcium indicator is required. By relying on previously published C. elegans data, it is difficult to know how differences in growth conditions or imaging conditions affect ASE responses. In addition, the paper would be strengthened by additional quantitative analysis of the calcium imaging data. For example, the paper states that 25 mM NH4Cl evokes a greater response in ASEL than 250 mM NH4Cl, but a quantitative comparison of the maximum responses to the two stimuli is not shown.

(6) It would be helpful to examine, or at least discuss, the other P. pacificus paralogs of Ce-gcy-22. Are they expressed in Ppa-ASER? How similar are the different paralogs? Additional discussion of the Ppa-gcy-22 gene expansion in P. pacificus would be especially helpful with respect to understanding the relatively minor phenotype of the Ppa-gcy-22.3 mutants.

(7) The calcium imaging data from Ppa-ASE is quite variable. It would be helpful to discuss this variability. It would also be helpful to clarify how the ASEL and ASER neurons are being conclusively identified during calcium imaging.

(8) More information about how the animals were treated prior to calcium imaging would be helpful. In particular, were they exposed to salt solutions prior to imaging? In addition, the animals are in an M9 buffer during imaging - does this affect calcium responses in Ppa-ASE and Ppa-AFD? More information about salt exposure, and how this affects neuron responses, would be very helpful.

(9) In Figure 6, the authors say that Ppa-gcy-22.3::GFP expression is absent in the Ppa-che-1(ot5012) mutant. However, based on the figure, it looks like there is some expression remaining. Is there a residual expression of Ppa-gcy-22.3::GFP in ASE or possibly ectopic expression in AFD? Does Ppa-che-1 regulate rGC expression in AFD? It would be helpful to address the role of Ppa-che-1 in AFD neuron differentiation.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

Mackie and colleagues compare chemosensory preferences between C. elegans and P. pacificus, and the cellular and molecular mechanisms underlying them. The nematodes have overlapping and distinct preferences for different salts. Although P. pacificus lacks the lsy-6 miRNA important for establishing asymmetry of the left/right ASE salt-sensing neurons in C. elegans, the authors find that P. pacificus ASE homologs achieve molecular (receptor expression) and functional (calcium response) asymmetry by alternative means. This work contributes an important comparison of how these two nematodes sense salts and highlights that evolution can find different ways to establish asymmetry in small nervous systems to optimize the processing of chemosensory cues in the environment.

Strengths:

The authors use clear and established methods to record the response of neurons to chemosensory cues. They were able to show clearly that ASEL/R are functionally asymmetric in P. pacificus, and combined with genetic perturbation establish a role for che-1-dependent gcy-22.3 in in the asymmetric response to NH4Cl.

Weaknesses:

The mechanism of lsy-6-independent establishment of ASEL/R asymmetry in P. pacificus remains uncharacterized.

We thank the reviewer for recognizing the novel contributions of our work in revealing the existence of alternative pathways for establishing neuronal lateral asymmetry without the lsy-6 miRNA in a divergent nematode species. We are certainly encouraged now to search for genetic factors that alter the exclusive asymmetric expression of gcy-22.3.

Reviewer #2 (Public review):

Summary:

In this manuscript, Mackie et al. investigate gustatory behavior and the neural basis of gustation in the predatory nematode Pristionchus pacificus. First, they show that the behavioral preferences of P. pacificus for gustatory cues differ from those reported for C. elegans. Next, they investigate the molecular mechanisms of salt sensing in P. pacificus. They show that although the C. elegans transcription factor gene che-1 is expressed specifically in the ASE neurons, the P. pacificus che-1 gene is expressed in the Ppa-ASE and Ppa-AFD neurons. Moreover, che-1 plays a less critical role in salt chemotaxis in P. pacificus than C. elegans. Chemogenetic silencing of Ppa-ASE and Ppa-AFD neurons results in more severe chemotaxis defects. The authors then use calcium imaging to show that both Ppa-ASE and Ppa-AFD neurons respond to salt stimuli. Calcium imaging experiments also reveal that the left and right Ppa-ASE neurons respond differently to salts, despite the fact that P. pacificus lacks lsy-6, a microRNA that is important for ASE left/right asymmetry in C. elegans. Finally, the authors show that the receptor guanylate cyclase gene Ppa-gcy-23.3 is expressed in the right Ppa-ASE neuron (Ppa-ASER) but not the left Ppa-ASE neuron (Ppa-ASEL) and is required for some of the gustatory responses of Ppa-ASER, further confirming that the Ppa-ASE neurons are asymmetric and suggesting that Ppa-GCY-23.3 is a gustatory receptor. Overall, this work provides insight into the evolution of gustation across nematode species. It illustrates how sensory neuron response properties and molecular mechanisms of cell fate determination can evolve to mediate species-specific behaviors. However, the paper would be greatly strengthened by a direct comparison of calcium responses to gustatory cues in C. elegans and P. pacificus, since the comparison currently relies entirely on published data for C. elegans, where the imaging parameters likely differ. In addition, the conclusions regarding Ppa-AFD neuron function would benefit from additional confirmation of AFD neuron identity. Finally, how prior salt exposure influences gustatory behavior and neural activity in P. pacificus is not discussed.

Strengths:

(1) This study provides exciting new insights into how gustatory behaviors and mechanisms differ in nematode species with different lifestyles and ecological niches. The results from salt chemotaxis experiments suggest that P. pacificus shows distinct gustatory preferences from C. elegans. Calcium imaging from Ppa-ASE neurons suggests that the response properties of the ASE neurons differ between the two species. In addition, an analysis of the expression and function of the transcription factor Ppa-che-1 reveals that mechanisms of ASE cell fate determination differ in C. elegans and P. pacificus, although the ASE neurons play a critical role in salt sensing in both species. Thus, the authors identify several differences in gustatory system development and function across nematode species.

(2) This is the first calcium imaging study of P. pacificus, and it offers some of the first insights into the evolution of gustatory neuron function across nematode species.

(3) This study addresses the mechanisms that lead to left/right asymmetry in nematodes. It reveals that the ASER and ASEL neurons differ in their response properties, but this asymmetry is achieved by molecular mechanisms that are at least partly distinct from those that operate in C. elegans. Notably, ASEL/R asymmetry in P. pacificus is achieved despite the lack of a P. pacificus lsy-6 homolog.

Weaknesses:

(1) The authors observe only weak attraction of C. elegans to NaCl. These results raise the question of whether the weak attraction observed is the result of the prior salt environment experienced by the worms. More generally, this study does not address how prior exposure to gustatory cues shapes gustatory responses in P. pacificus. Is salt sensing in P. pacificus subject to the same type of experience-dependent modulation as salt sensing in C. elegans?

We tested if starving animals in the presence of a certain salt will result in those animals avoiding it. However, under our experimental conditions we were unable to detect experiencedependent modulation either in P. pacificus or in C. elegans.

(2) A key finding of this paper is that the Ppa-CHE-1 transcription factor is expressed in the PpaAFD neurons as well as the Ppa-ASE neurons, despite the fact that Ce-CHE-1 is expressed specifically in Ce-ASE. However, additional verification of Ppa-AFD neuron identity is required. Based on the image shown in the manuscript, it is difficult to unequivocally identify the second pair of CHE-1-positive head neurons as the Ppa-AFD neurons. Ppa-AFD neuron identity could be verified by confocal imaging of the CHE-1-positive neurons, co-expression of Ppa-che1p::GFP with a likely AFD reporter, thermotaxis assays with Ppa-che-1 mutants, and/or calcium imaging from the putative Ppa-AFD neurons.

In the revised manuscript, we provide additional and, we believe, conclusive evidence for our correct identification of Ppa-AFD neuron being another CHE-1 expressing neuron. Specifically, we have constructed and characterized 2 independent reporter strains of Ppa-ttx-1, a putative homolog of the AFD terminal selector in C. elegans. There are two pairs of ttx-1p::rfp expressing amphid neurons. The anterior neuronal pair have finger-like endings that are unique for AFD neurons compared to the dendritic endings of the 11 other amphid neuron pairs (no neuron type has a wing morphology in P. pacificus). Their cell bodies are detected in the newly tagged TTX-1::ALFA strain that co-localize with the anterior pair of che-1::gfp-expressing amphid neurons (n=15, J2-Adult).

We note that the identity of the posterior pair of amphid neurons differs between the ttx-1p::rfp promoter fusion reporter and TTX-1::ALFA strains– the ttx-1p::rfp posterior amphid pair overlaps with the gcy-22.3p::gfp reporter (ASER) but the TTX-1::ALFA posterior amphid pair do not overlap with the posterior pair of che-1::gfp-expressing amphid neurons (n=15). Given that there are 4 splice forms detected by RNAseq (Transcriptome Assembly Trinity, 2016; www.pristionchus.org), this discrepancy between the Ppa-ttx-1 promoter fusion reporter and the endogenous expression of the Ppa-TTX-1 C-terminally tagged to the only splice form containing Exon 18 (ppa_stranded_DN30925_c0_g1_i5, the most 3’ exon) may be due to differential expression of different splice variants in AFD, ASE, and another unidentified amphid neuron types.

Although we also made reporter strains of two putative AFD markers, Ppa-gcy-8.1 (PPA24212)p::gfp; csuEx101 and Ppa-gcy-8.2 (PPA41407)p::gfp; csuEx100, neither reporter showed neuronal expression.

(3) Loss of Ppa-che-1 causes a less severe phenotype than loss of Ce-che-1. However, the loss of Ppa-che-1::RFP expression in ASE but not AFD raises the question of whether there might be additional start sites in the Ppa-che-1 gene downstream of the mutation sites. It would be helpful to know whether there are multiple isoforms of Ppa-che-1, and if so, whether the exon with the introduced frameshift is present in all isoforms and results in complete loss of Ppa-CHE-1 protein.

According to www.pristionchus.org (Transcriptome Assembly Trinity), there is only a single detectable splice form by RNAseq. Once we have a Ppa-AFD-specific marker, we would be able to determine how much of the AFD terminal effector identify (e.g. expression of gcy-8 paralogs) is effected by the loss of Ppa-che-1 function.

(4) The authors show that silencing Ppa-ASE has a dramatic effect on salt chemotaxis behavior. However, these data lack control with histamine-treated wild-type animals, with the result that the phenotype of Ppa-ASE-silenced animals could result from exposure to histamine dihydrochloride. This is an especially important control in the context of salt sensing, where histamine dihydrochloride could alter behavioral responses to other salts.

We have inadvertently left out this important control. Because the HisCl1 transgene is on a randomly segregating transgene array, we have scored worms with and without the transgene expressing the co-injection marker (Ppa-egl-20p::rfp, a marker in the tail) to show that the presence of the transgene is necessary for the histamine-dependent knockdown of NH4Br attraction. This control is added as Figure S2.

(5) The calcium imaging data in the paper suggest that the Ppa-ASE and Ce-ASE neurons respond differently to salt solutions. However, to make this point, a direct comparison of calcium responses in C. elegans and P. pacificus using the same calcium indicator is required. By relying on previously published C. elegans data, it is difficult to know how differences in growth conditions or imaging conditions affect ASE responses. In addition, the paper would be strengthened by additional quantitative analysis of the calcium imaging data. For example, the paper states that 25 mM NH4Cl evokes a greater response in ASEL than 250 mM NH4Cl, but a quantitative comparison of the maximum responses to the two stimuli is not shown.

We understand that side-by-side comparisons with C. elegans using the same calcium indicator would lend more credence to the differences we observed in P. pacificus versus published findings in C. elegans from the past decades, but are not currently in a position to conduct these experiments in parallel.

(6) It would be helpful to examine, or at least discuss, the other P. pacificus paralogs of Ce-gcy22. Are they expressed in Ppa-ASER? How similar are the different paralogs? Additional discussion of the Ppa-gcy-22 gene expansion in P. pacificus would be especially helpful with respect to understanding the relatively minor phenotype of the Ppa-gcy-22.3 mutants.

In P. pacificus, there are 5 gcy-22-like paralogs and 3 gcy-7-like paralogs, which together form a subclade that is clearly distinct from the 1-1 Cel-gcy-22, Cel-gcy-5, and Cel-gcy-7 orthologs in a phylogenetic tree containing all rGCs in P. pacificus, C. elegans, and C. briggssae (Hong et al, eLife, 2019). In Ortiz et al (2006 and 2009), Cel-gcy-22 stands out from other ASER-type gcy genes (gcy-1, gcy-4, gcy-5) in being located on a separate chromosome (Chr. V) as well as in having a wider range of defects in chemoattraction towards salt ions. Given that the 5 P. pacificus gcy-22-like paralogs are located on 3 separate chromosomes without clear synteny to their C. elegans counterparts, it is likely that the gcy-22 paralogs emerged from independent and repeated gene duplication events after the separation of these Caenorhabditis and Pristionchus lineages. Our reporter strains for two other P. pacificus gcy-22-like paralogs either did not exhibit expression in amphid neurons (Ppa-gcy-22.1p::GFP, ) or exhibited expression in multiple neuron types in addition to a putative ASE neuron (Ppa-gcy-22.4p::GFP). We have expanded the discussion on the other P. pacificus gcy-22 paralogs.

(7) The calcium imaging data from Ppa-ASE is quite variable. It would be helpful to discuss this variability. It would also be helpful to clarify how the ASEL and ASER neurons are being conclusively identified during calcium imaging.

For each animal, the orientation of the nose and vulva were recorded and used as a guide to determine the ventral and dorsal sides of the worm, and subsequently, the left and right sides of the worm. Accounting for the plane of focus of the neuron pairs as viewed through the microscope, it was then determined whether the imaged neuron was the worm’s left or right neuron of each pair. We added this explanation to the Methods.

(8) More information about how the animals were treated prior to calcium imaging would be helpful. In particular, were they exposed to salt solutions prior to imaging? In addition, the animals are in an M9 buffer during imaging - does this affect calcium responses in Ppa-ASE and Ppa-AFD? More information about salt exposure, and how this affects neuron responses, would be very helpful.

Prior to calcium imaging, animals were picked from their cultivation plates (using an eyelash pick to minimize bacteria transfer) and placed in loading solution (M9 buffer with 0.1% Tween20 and 1.5 mM tetramisole hydrochloride, as indicated in the Method) to immobilize the animals until they were visibly completely immobilized.

(9) In Figure 6, the authors say that Ppa-gcy-22.3::GFP expression is absent in the Ppa-che1(ot5012) mutant. However, based on the figure, it looks like there is some expression remaining. Is there a residual expression of Ppa-gcy-22.3::GFP in ASE or possibly ectopic expression in AFD? Does Ppa-che-1 regulate rGC expression in AFD? It would be helpful to address the role of Ppa-che-1 in AFD neuron differentiation.

In Figure 6C, the green signal is autofluorescence in the gut, and there is no GFP expression detected in any of the 55 che-1(-) animals we examined. We are currently developing AFDspecific rGC markers (gcy-8 homologs) to be able to examine the role of Ppa-CHE-1 in regulating AFD identity.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Abstract: 'how does sensory diversity prevail within this neuronal constraint?' - could be clearer as 'numerical constraint' or 'neuron number constraint'.

We have clarified this passage as ‘…constraint in neuron number’.

(2) 'Sensory neurons in the Pristionchus pacificus' - should get rid of the 'the'.

We have removed the ‘the’.

(3) Figure 2: We have had some good results with the ALFA tag using a similar approach (tagging endogenous loci using CRISPR). I'm not sure if it is a Pristionchus thing, or if it is a result of our different protocols, but our staining appears stronger with less background. We use an adaptation of the Finney-Ruvkin protocol, which includes MeOH in the primary fixation with PFA, and overcomes the cuticle barrier with some LN2 cracking, DTT, then H2O2. No collagenase. If you haven't tested it already it might be worth comparing the next time you have a need for immunostaining.

We appreciate this suggestion. Our staining protocol uses paraformaldehyde fixation. We observed consistent and clear staining in only 4 neurons in CHE-1::ALFA animals but more background signals from TTX-1::ALFA in Figure 2I-J in that could benefit from improved immunostaining protocol.

(4) Page 6: 'By crossing the che-1 reporter transgene into a che-1 mutant background (see below), we also found that che-1 autoregulates its own expression (Figure 2F), as it does in C. elegans' - it took me some effort to understand this. It might make it easier for future readers if this is explained more clearly.

We understand this confusion and have changed the wording along with a supporting table with a more detailed account of che-1p::RFP expression in both ASE and AFD neurons in wildtype and che-1(-) backgrounds in the Results.

(5) Line numbers would make it easier for reviewers to reference the text.

We have added line numbers.

(6) Page 7: is 250mM NH4Cl an ecologically relevant concentration? When does off-target/nonspecific activation of odorant receptors become an issue? Some discussion of this could help readers assess the relevance of the salt concentrations used.

This is a great question but one that is difficult to reconcile between experimental conditions that often use 2.5M salt as point-source to establish salt gradients versus ecologically relevant concentrations that are very heterogenous in salinity. Efforts to show C. elegans can tolerate similar levels of salinity between 0.20-0.30 M without adverse effects have been recorded previously (Hu et al., Analytica Chimica Acta 2015; Mah et al. Expedition 2017).

(7) It would be nice for readers to have a short orientation to the ecological relevance of the different salts - e.g. why Pristionchus has a particular taste for ammonium salts.

Pristionchus species are entomophilic and most frequently found to be associated with beetles in a necromenic manner. Insect cadavers could thus represent sources of ammonium in the soil. Additionally, ammonium salts could represent a biological signature of other nematodes that the predatory morphs of P. pacificus could interpret as prey. We have added the possible ecological relevance of ammonium salts into the Discussion.

(8) Page 11: 'multiple P. pacificus che-1p::GCaMP strains did not exhibit sufficient basal fluorescence to allow for image tracking and direct comparison'. 500ms exposure to get enough signal from RCaMP is slow, but based on the figures it still seems enough to capture things. If image tracking was the issue, then using GCaMP6s with SL2-RFP or similar in conjunction with a beam splitter enables tracking when the GCaMP signal is low. Might be an option for the future.

These are very helpful suggestions and we hope to eventually develop an improved che1p::GCaMP strain for future studies.

(9) Sometimes C. elegans genes are referred to as 'C. elegans [gene name]' and sometimes 'Cel [gene name]'. Should be consistent. Same with Pristionchus.

We have now combed through and corrected the inconsistencies in nomenclature.

(10) Pg 12 - '...supports the likelihood that AFD receives inputs, possibly neuropeptidergic, from other amphid neurons' - the neuropeptidergic part could do with some justification.

Because the AFD neurons are not exposed directly to the environment through the amphid channel like the ASE and other amphid neurons, the calcium responses to salts detected in the AFD likely originate from sensory neurons connected to the AFD. However, because there is no synaptic connection from other amphid neurons to the AFD neurons in P. pacificus (unlike in C. elegans; Hong et al, eLife, 2019), it is likely that neuropeptides connect other sensory neurons to the AFDs. To avoid unnecessary confusion, we have removed “possibly neuropeptidergic.”

(11) Pg16: the link to the Hallam lab codon adaptor has a space in the middle. Also, the paper should be cited along with the web address (Bryant and Hallam, 2021).

We have now added the proper link, plus in-text citation. https://hallemlab.shinyapps.io/Wild_Worm_Codon_Adapter/ (Bryant and Hallem, 2021)

Full citation:

Astra S Bryant, Elissa A Hallem, The Wild Worm Codon Adapter: a web tool for automated codon adaptation of transgenes for expression in non-Caenorhabditis nematodes, G3 Genes|Genomes|Genetics, Volume 11, Issue 7, July 2021, jkab146, https://doi.org/10.1093/g3journal/jkab146

Reviewer #2 (Recommendations for the authors):

(1) In Figure 1, the legend states that the population tested was "J4/L4 larvae and young adult hermaphrodites," whereas in the main text, the population was described as "adult hermaphrodites." Please clarify which ages were tested.

We have tested J4-Adult stage hermaphrodites and have made the appropriate corrections in the text.

(2) The authors state that "in contrast to C. elegans, we find that P. pacificus is only moderately and weakly attracted to NaCl and LiCl, respectively." However, this statement does not reflect the data shown in Figure 1, where there is no significant difference between C. elegans and P. pacificus - both species show at most weak attraction to NaCl.

Although there is no statistically significant difference in NaCl attraction between P. pacificus and C. elegans, NaCl attraction in P. pacificus is significantly lower than its attraction to all 3 ammonium salts when compared to C. elegans. We have rephrased this statement as relative differences in the Results and updated the Figure legend.

(3) In Figure 1, the comparisons between C. elegans and P. pacificus should be made using a two-way ANOVA rather than multiple t-tests. Also, the sample sizes should be stated (so the reader does not need to count the circles) and the error bars should be defined.

We performed the 2-way ANOVA to detect differences between C. elegans and P. pacificus for the same salt and between salts within each species. We also indicated the sample size on the figure and defined the error bars.

Significance:

For comparisons of different salt responses within the same species:

- For C. elegans, NH4Br vs NH4Cl (**p<0.01), NH4Cl vs NH4I (* p<0.05), and NH4Cl vs NaCl (* p<0.05). All other comparisons are not significant.

- For P. pacificus, all salts showed (****p<0.0001) when compared to NaAc and to NH4Ac, except for NH4Ac and NaAc compared to each other (ns). Also, NH4Cl showed (*p<0.05) and NH4I showed (***p<0.001) when compared with LiCl and NaCl. All other comparisons are not significant.

For comparisons of salt responses between different species (N2 vs PS312):

- NH4I and LiCl (*p<0.05); NaAc and NH4Ac (****p<0.0001)

(4) It might be worth doing a power analysis on the data in Figure 3B. If the data are underpowered, this might explain why there is a difference in NH4Br response with one of the null mutants but not the other.

For responses to NH4Cl, since both che-1 mutants (rather than just one) showed significant difference compared to wildtype, we conducted a power analysis based on the effect size of that difference (~1.2; large). Given this effect size, the sample size for future experiments should be 12 (ANOVA).

For responses to NH4Br and given the effect size of the difference seen between wildtype (PS312) and ot5012 (~0.8; large), the sample size for future experiments should be 18 (ANOVA) for a power value of 0.8. Therefore, it is possible that the sample size of 12 for the current experiment was too small to detect a possible difference between the ot5013 alleles and wildtype.

(5) It would be helpful to discuss why silencing Ppa-ASE might result in a switch from attractive to repulsive responses to some of the tested gustatory cues.

For similar assays using Ppa-odr-3p::HisCl1, increasing histamine concentration led to decreasing C.I. for a given odorant (myristate, a P. pacificus-specific attractant). It is likely that the amount of histamine treatment for knockdown to zero (i.e. without a valence change) will differ depending on the attractant.

(6) The statistical tests used in Figure 3 are not stated.

Figure 3 used Two-way ANOVA with Dunnett’s post hoc test. We have now added the test in the figure legend.

(7) It would be helpful to examine the responses of ASER to the full salt panel in the Ppa-gcy-22.3 vs. wild-type backgrounds.

We understand that future experiments examining neuron responses to the full salt panel for wildtype and gcy-22.3 mutants would provide further information about the salts and specific ions associated with the GCY-22.3 receptor. However, we have tested a broader range of salts (although not yet the full panel) for behavioral assays in wildtype vs gcy-22.3 mutants, which we have included as part of an added Figure 8.

(8) The controls shown in Figure S1 may not be adequate. Ideally, the same sample size would be used for the control, allowing differences between control worms and experimental worms to be quantified.

Although we had not conducted an equal number of negative controls using green light without salt stimuli due to resource constraints (6 control vs ~10-19 test), we provided individual recordings with stimuli to show that conditions we interpreted as having responses rarely showed responses resembling the negative controls. Similarly, those we interpreted as having no responses to stimuli mostly resembled the no-stimuli controls (e.g. WT to 25 mM NH4Cl, gcy22.3 mutant to 250 mM NH4Cl).

(9) An osmolarity control would be helpful for the calcium imaging experiments.

We acknowledge that future calcium imaging experiments featuring different salt concentrations could benefit from osmolarity controls.

(10) In Figure S7, more information about the microfluidic chip design is needed.

The chip design features a U-shaped worm trap to facilitate loading the worm head-first, with a tapered opening to ensure the worm fits snugly and will not slide too far forward during recording. The outer two chip channels hold buffer solution and can be switched open (ON) or closed (OFF) by the Valvebank. The inner two chip channels hold experimental solutions. The inner channel closer to the worm trap holds the control solution, and the inner channel farther from the worm trap holds the stimulant solution.

We have added an image of the chip in Figure S7 and further description in the legend.

(11) Throughout the manuscript, the discussion of the salt stimuli focuses on the salts more than the ions. More discussion of which ions are eliciting responses (both behavioral and neuronal responses) would be helpful.

In Figure 7, the gcy-22.3 defect resulted in a statistically significant reduction in response only towards NH4Cl but not towards NaCl, which suggests ASER is the primary neuron detecting NH4+ ions. To extend the description of the gcy-22.3 mutant defects to other ions, we have added a Figure 8: chemotaxis on various salt backgrounds. We found only a mild increase in attraction towards NH4+ by both gcy-22.3 mutant alleles, but wild-type in their responses toward Cl-, Na+, or I-. The switch in the direction of change between the behavioral (enhanced) and calcium imaging result (reduced) suggests the behavioral response to ammonium ions likely involves additional receptors and neurons.

Minor comments:

(1) The full species name of "C. elegans" should be written out upon first use.

We have added ‘Caenorhabditis elegans’ to its first mention.

(2) In the legend of Figure 1, "N2" should not be in italics.

We have made the correction.

(3) The "che-1" gene should be in lowercase, even when it is at the start of the sentence.

We have made the correction.

(4) Throughout the manuscript, "HisCl" should be "HisCl1."

We have made these corrections to ‘HisCl1’.

(5) Figure 3A would benefit from more context, such as the format seen in Figure 7A. It would also help to have more information in the legend (e.g., blue boxes are exons, etc.).

(6) "Since NH4I sensation is affected by silencing of che-1(+) neurons but is unaffected in che-1 mutants, ASE differentiation may be more greatly impacted by the silencing of ASE than by the loss of che-1": I don't think this is exactly what the authors mean. I would say, "ASE function may be more greatly impacted...".

We have changed ‘differentiation’ to ‘function’ in this passage.

(7) In Figure 7F-G, the AFD neurons are referred to as AFD in the figure title but AM12 in the graph. This is confusing.

Thank you for noticing this oversight. We have corrected “AM12” to “AFD”.

(8) In Figure 7, the legend suggests that comparisons within the same genotype were analyzed. I do not see these comparisons in the figure. In which cases were comparisons within the same genotype made?

Correct, we performed additional tests between ON and OFF states within the same genotypes (WT and mutant) but did not find significant differences. To avoid unnecessary confusion, we have removed this sentence.

(9) The nomenclature used for the transgenic animals is unconventional. For example, normally the calcium imaging line would be listed as csuEx93[Ppa-che-1p::optRCaMP] instead of Ppache-1p::optRCaMP(csuEx93).

We have made these corrections to the nomenclature.

(10) Figure S6 appears to come out of order. Also, it would be nice to have more of a legend for this figure. The format of the figure could also be improved for clarity.

We have corrected Figure S6 (now S8) and added more information to the legend.

(11) Methods section, Chemotaxis assays: "Most assays lasted ~3.5 hours at room temperature in line with the speed of P. pacificus without food..." It's not clear what this means. Does it take the worms 3.5 hours to crawl across the surface of the plate?

Correct, P. pacificus requires 3-4 hours to crawl across the surface of the plate, which is the standard time for chemotaxis assays for some odors and all salts. We have added this clarification to the Methods.

Version 2

Abstract

Abstract

Animals with small nervous systems have a limited number of sensory neurons that must encode information from a changing environment. This problem is particularly exacerbated in nematodes that populate a wide variety of distinct ecological niches but only have a few sensory neurons available to encode multiple modalities. How does sensory diversity prevail within this constraint in neuron number? To identify the genetic basis for patterning different nervous systems, we demonstrate that sensory neurons in Pristionchus pacificus respond to various salt sensory cues in a manner that is partially distinct from that of the distantly related nematode Caenorhabditis elegans. By visualizing neuronal activity patterns, we show that contrary to previous expectations based on its genome sequence, the salt responses of P. pacificus are encoded in a left/right asymmetric manner in the bilateral ASE neuron pair. Our study illustrates patterns of evolutionary stability and change in the gustatory system of nematodes.

Animals with small nervous systems have evolved left and right asymmetry in their neurons to process various salts at different concentrations.

Introduction

Nematodes form a vast array of ecological relationships, from specialized parasite-host dependencies to nematode-microbial interactions, each one demanding exquisitely fine-tuned sets of sensory palates that span multiple modalities (1–8). Yet, the number of sensory neurons across diverse nematode species seems to be constrained (9–11). Several free-living and parasitic nematode species examined by serial section electron microscopy have nearly identical numbers of 12-13 pairs of head sensory neurons, known as the amphid neurons (11–15). How does sensory diversity arise within this constraint in neuron number? When coupled with well-described neuronal anatomy, this conserved neuron count allows for detailed comparisons at the single-cell resolution and represents an opportunity to interrogate how sensory cues are processed by anatomically similar nervous systems to produce species-specific or developmental stage-dependent behavioral outputs. To identify the genetic basis for patterning different nervous systems and to understand the processes that underlie evolutionary changes in adapting to different environments, several comparative model systems have been developed to promote comparisons to the well-studied nematode Caenorhabditis elegans at the genetic and cellular levels, including the predatory entomophilic nematode, Pristionchus pacificus (11, 16–19).

As expected from their association with insects in the wild, the olfactory preferences of P. pacificus are distinct from those of C. elegans and the human parasite Strongyloides stercoralis (2, 3, 20), but little is known about P. pacificus responses to water-soluble compounds. In C. elegans, the main salt receptor neuron class comprises of a bilateral pair of left and right ASE neurons (ASEL and ASER), which serve to induce an attractive locomotory response toward an increase in salt concentration (21). The gene che-1 (chemosensory defective) encodes a transcription factor that is exclusively expressed in the ASE neurons and is required for their proper differentiation, such that a che-1 mutant results in defective salt attraction (22–25).

One major role of C. elegans CHE-1 is to promote lateral asymmetry in the ASE neurons. The left and right ASE neurons asymmetrically express receptor-type guanylyl cylases (rGCs, encoded by gcy genes) (26). This finding led to the realization that the ASE neurons are lateralized, such that the left and the right ASE neurons differentially respond to distinct salt ions (27–30). This observation led in turn to the identification of a complex gene regulatory network that genetically programs the distinct sensory potentials of the left and right ASE neurons (31), which includes a miRNA, lsy-6, at the top of this gene regulatory network (31–33). However, the lsy-6 miRNA evolved selectively in the Caenorhabditis genus (34) and is absent in P. pacificus. Moreover, P. pacificus does not show an expansion of the ASEL-type and ASER-type rGCs, as C. elegans does (11) . With these two genomic observations in mind, we had previously proposed that the ASE neurons are unlikely to be lateralized in P. pacificus (11).

However, we now revise this view in light of our work on mapping chemosensory responses in P. pacificus on the level of behavior and neuronal activity. We demonstrate that P. pacificus does in fact show lateralized chemosensory profiles, indicating that P. pacificus must have evolved independent means to establish ASE laterality. We also show that the tastant palate of P. pacificus is distinct from that of C. elegans, and that its dependence on ASE, as well as its terminal selector transcription factor che-1, have also diverged.

Results

P. pacificus and C. elegans differ in their behavioral responses to salts

Previous cross-species comparisons between P. pacificus and C. elegans indicated strong differences in olfaction preferences that reflect their divergent evolutionary histories and ecology (2). To identify the neurons that mediate gustation, we first compared the chemosensory profiles of these two species toward water-soluble ions. In this survey, we found ammonium salts to be the strongest attractants to wildtype P. pacificus J4 to adult hermaphrodites (NH4Br, NH4Cl, NH4I), with NH4I significantly more attractive to P. pacificus compared to C. elegans (Fig. 1). Notably, in contrast to C. elegans (35), we find that P. pacificus is less attracted to NaCl and LiCl compared to the ammonium salts (Fig. 1). Also, P. pacificus is repulsed by acetate salts (NaAc, NH4Ac), which induce attractive responses in C. elegans (36). We conclude that P. pacificus and C. elegans display differences in their salt preferences.

A comparison of chemotaxis responses to water-soluble ions between P. pacificus and C. elegans.

J4 to Adult hermaphrodites from the two species responded to NH4I, LiCl, and acetates significantly differently. Using two-way ANOVA, significant difference found between wildtype P. pacificus and C. elegans for the same salt is indicated above each pair (*p<0.05, ****p<0.0001), while the differences within P. pacificus is as follows: all salts showed difference when compared to NaAc and to NH4Ac (****p<0.0001), but not between NH4Ac and NaAc. Both LiCl and NaCl attraction are significantly lower than NH4Cl (*p<0.05) and NH4I (***p<0.001). Error bars denote standard error of the mean and the sample sizes are indicated on the bottom.

Ppa-che-1 shows similarities and difference to Cel-che-1 in both expression and function

The C. elegans che-1 mutant (chemotaxis-defective) was originally isolated for its inability to respond to a broad panel of salt tastants (22, 35), including those described above (29, 30, 36). che-1 was found to encode for a Zn finger transcription factor that is exclusively expressed in the ASE neuron pair (25), which through laser ablations had been found to be the main salt receptor neurons (37). che-1 was found to control the entire differentiation of the ASE neurons, including the expression of putative receptors of the GCY receptor guanylyl cyclase family (23–25, 38).

To assess whether che-1 performs a similar function in salt perception for P. pacificus as for C. elegans, we analyzed the expression and function of the Ppa-che-1 ortholog. In a previous paper, we reported that the 5’ region of the sole Pristionchus che-1 ortholog directs reporter expression to the Ppa ASE and Ppa ASG neuron classes (11). Re-examination of our provisional cell identifications using a newly-generated che-1p::GFP strain with stronger neurite expression revealed highly elaborated finger-like dendritic endings in the more anterior amphid neuron that could unambiguously be assigned to the AFD neurons (Fig. 2A-C), prompting us to reassign expression of che-1 to ASE and AFD.

P. pacificus che-1 expression in ASE and AFD amphid neurons.

(A-B) The che-1::GFP marker in the che-1::RCaMP reporter strain is expressed in the ASE and AFD amphid neurons. (C) che-1::GFP expression is detectable in the morphologically distinct AFD neurons with ‘finger’-like dendritic endings. (D-E) Immunostaining of CHE-1::ALFA shows two pairs of amphid neuron cell bodies corresponding to the ASE and AFD neurons; dorsal-ventral view (n=114). (F) The loss of che-1 results in loss che-1::RFP expression in the ASE (circle) while retaining reduced AFD expression. (G) ttx-1p::RFP expression in the AFD neurons with ‘finger’-like dendritic endings. Inset shows expanded inverted black-white image of the AFD dendritic ending. (H) AFD expression of the same ttx-1::RFP animal shown in (G) in a different plane with cell body in focus. (I-J) Immunostaining of TTX-1::ALFA (red) shows one pair of amphid neuron cell bodies co-localizing with the anterior pair of che-1::GFP-expressing AFD neurons (yellow); dorsal-ventral view (n=13). Anterior is left and the scale bar in (C) represents 50 µm for all panels except for the G inset, which is 5 µm.

We confirmed that the che-1 reporter transgene indicates the full expression of the endogenous che-1 locus by tagging the endogenous che-1 locus with an ALFA-tag (39). The che-1::ALFA animals showed staining in 2 pairs of head neurons whose position is consistent with being the ASE and AFD neurons (Fig. 2D-E). By crossing the che-1 reporter transgene into a che-1 mutant background (see below), we also found that che-1 autoregulates its own expression especially in the ASE neurons (Fig. 2F, GFP expression in 97% in WT (n= 36) vs 4% in che-1(-)(n=48)), as it does in C. elegans (25). The reduction of che-1p::GFP in the AFDs was also observed (GFP expression in 100% in WT (n= 36) vs 33% in che-1(-)(n=48)).

To provide further evidence that Ppa-CHE-1 is indeed expressed in the Ppa-AFD neurons, we analyzed the expression of the Ppa-ttx-1p::RFP reporter. In C. elegans, the OTX homeodomain transcription factor TTX-1 is a terminal selector expressed in the AFD neurons required for designating the AFD fate (Satterley 2001). We found ttx-1p::GFP to be strongly expressed in a pair of neurons with the hallmark finger-like dendritic ending of the AFD neurons (Fig. 2G-H), as well as expression in other head and tail neurons (possibly RIP and RIB) and cells likely to be the pharyngeal marginal cells based on likely conservation with Cel-ttx-1 expression (Reilly 2022). To confirm that endogenous TTX-1 protein expression is co-expressed with che-1p::GFP, we examined C-terminally tagged ttx-1:ALFA animals and found co-localization in the same anterior pair of amphid neurons, but no co-expression in the posterior pair of amphid neurons (Fig. 2I-J, n=15). Interestingly, the promoter expression in the posterior pair of amphid neurons in animals with both ttx-1p::RFP and gcy-22.3p::GFP reporters do co-localize in the ASER neurons (Fig. S1), possibly due to differences between the expression patterns of the 4 possible ttx-1 splice forms. The transcriptional and protein co-expression of ttx-1 and che-1 in the AFD neurons unequivocally show the expression of che-1 in both AFD and ASE neurons in P. pacificus.

Next, we examined whether P. pacificus salt responsiveness shows similar che-1 dependence as in C. elegans. We generated two putative null alleles in the Ppa-che-1 homolog using CRISPR/Cas9 genome engineering, through introduction of small deletions in the first exon of the gene, thereby resulting in frameshift and premature stops (Fig. 3A). Both Ppa-che-1 alleles exhibited defects in attraction towards NH4Cl and LiCl compared to wildtype (Fig. 3B). However, unlike in C. elegans, Ppa-che-1 mutants showed no detectable difference in responses to NH4I, NaCl, and NaAc.

che-1 expressing amphid neurons are required for sensing water-soluble ions in P. pacificus.

(A) The che-1 locus with CRISPR/Cas9-induced mutations in Exon 1. (B) The che-1 mutants show defects in attraction towards NH4Br, NH4Cl, and LiCl. Sample sizes are indicated below for attractants and above for repellent. (C) The che-1p::HisCl1 animals lose attraction towards NH4Br, NH4Cl, and NH4I in a histamine dependent manner. Sample sizes are indicated at the base of each bar. (D) A schematic map of the P. pacificus AM7(ASE) and AM12(AFD) amphid neurons that express the che-1p::RCaMP used in calcium imaging. The ASE axons are the only amphid axons in P. pacificus to cross each other over the dorsal lateral midline contralaterally. **P<0.01, *P<0.05 Two-way ANOVA with Dunnett’s post-hoc comparison showing alleles with significant difference to wildtype P. pacificus (PS312). ****P<0.0001 Two-way ANOVA showing significant difference between water control and histamine treatment.

We considered two different possibilities for the behavioral differences of Ppa-che-1 and Cel-che-1 mutants. P. pacificus may use sensory neurons other than ASE to sense these cues, or alternatively, Ppa-che-1 may not have the same fundamental impact on ASE function in P. pacificus as it does in C. elegans. To explore these different possibilities, we silenced che-1 expressing neurons by expressing codon-optimized HisCl1 channel under control of the che-1 promoter. Histamine-treated che-1p::HisCl1 animals showed complete loss of attraction to NH4Br, NH4Cl, and NH4I but did not significantly alter their repulsive response to NaAc (Fig. 3C). As a control, we show that the presence of the che-1p::HisCl1 transgene was necessary for the knockdown of NH4Br attraction (Fig. S2). These findings corroborate that NaAc is sensed by neurons other than ASE (or AFD, in which the che-1 promoter also drives HisCl1). Since NH4I sensation is affected by silencing of che-1(+) neurons but is unaffected in che-1 mutants, ASE function may be more greatly impacted by the silencing of ASE than by the loss of che-1.

The ASE neurons show left/right asymmetric responses to salt

To assess whether Ppa ASE neurons show the same lateralized response to salt ions as Cel ASE neurons, we generated transgenic P. pacificus lines that express RCaMP in ASE neurons and assessed calcium responses to attractive salts (Fig. 4, S3-S4). Specifically, we looked for changes in calcium levels immediately after the addition and removal of specific salts. When 250 mM NH4Cl is administered, an ‘ON’ response is observed as calcium transiently increases in the left ASE neuron (ASEL). In contrast, an ‘OFF’ response was observed as calcium sharply dips in the right ASE neuron (ASER) before quickly returning to baseline when the salt was removed, which suggests hyperpolarization of this neuron (Fig. 4A-B)(29). However, when presented with a 10-fold lower concentration of 25 mM NH4Cl, the ‘OFF’ response completely disappeared in ASER while the ‘ON’ response became more pronounced in ASEL. The ASER responses to 250 mM and 25 mM NaCl (Fig. 4D) were very similar to the ‘OFF’ response (including hyperpolarization) observed for NH4Cl, but the ASEL responses differ between the two salts: instead of the ‘ON’ response expected of ASEL, we observed a relatively weak ‘OFF’ response without the characteristic hyperpolarization but accompanied by an attenuated ‘bump’, a profile that we classify as an ‘OFF-2’ response (Fig. 4C). Finally, we examined the response to NH4I and found that it also elicited laterally asymmetric responses, but yet again distinct from the responses to both NH4Cl and NaCl, with ASEL showing a strong ‘ON’ response, and ASER showing an ‘ON-OFF’ biphasic response to 250 mM NH4I (Fig. 4E-F)(40, 41). Interestingly, the ASER exhibited an ‘OFF’-only response to 25 mM NH4I, which was not observed for the same concentration of NH4Cl and NaCl, and thus may reflect the higher response to NH4I than to NH4Cl and NaCl in the behavior assays. Altogether, P. pacificus ASE neurons clearly show left/right asymmetric responses to salt attractants and these asymmetric responses show similarities and differences to the C. elegans ASE taste neurons (see Discussion).

ASEL and ASER responses to different concentrations of NH4Cl, NaCl, and NH4I.

Average percent change in RCaMP fluorescence (dF/F) over time (seconds) of tracked ASE (left and right) sensory neurons in P. pacificus. Salts were presented at 10s (“ON”, left vertical line) for a duration of 20s, and then removed (“OFF”, right vertical line) for the remaining 30s; the total recording time was 60s. (A-B) ASEL and ASER neuron responses to high (250 mM, red) compared to low (25 mM, blue) concentrations of NH4Cl. (C-D) ASEL and ASER neuron responses to NaCl. (E-F) ASEL and ASER neuron responses to NH4I. Shaded ribbons represent 95% confidence intervals. Shaded ribbons in (E-F) have been cropped to maintain consistent y-axes across the plots, allowing for easier comparison. Sample sizes are indicated (n).

The AFD neurons also respond to salts in P. pacificus

The RCaMP line that we used to assess calcium responses in ASE is also expressed in AFD, allowing us to simultaneously examine the calcium responses in the AFD (AM12) neurons (Fig. 5, S1, S5-S7). Surprisingly, we detected a distinctly ‘ON-OFF’ biphasic response to all 3 salt types at both concentrations. Specifically, although we observed weaker or comparable responses in AFD neurons when compared to either ASE neuron’s response towards 250 mM NH4Cl, NaCl and NH4I (Fig. 5A-B, 5E-F, 5I-J), the AFD responses were more robust than ASER toward 25 mM NH4Cl, NaCl, and NH4I (Fig. 5D, 5H, 5L). Specifically, AFD neurons responded strongly to 25 mM NaCl when neither one of the ASE neurons showed a positive response (Fig. 5G-H). Averaging the calcium transients separately by AFD left versus right did not result in significant differences in the shape of the neuronal calcium responses, with the exception of the AFDR responses to higher versus lower concentrations of NH4I (Fig. S6-S7). We have not further pursued whether these AFD responses are a reflection of a direct perception of salt or a secondary consequence of communication of salt-perceptive neurons (like ASE) to AFD.

Combined AFD responses in comparison to ASE left or right neuron responses to NH4Cl, NaCl, and NH4I.

Average percent change in RCaMP fluorescence (dF/F) over time (seconds) as described in Fig. 4. Averaged combined AFD (left and right neurons, orange) compared to left or right ASE (green) responses to (A-B) 250mM NH4Cl (C-D) 25mM NH4Cl (E-F) 250mM NaCl (G-H) 25mM NaCl (I-J) 250mM NH4I (K-L) 25mM NH4I. Shaded ribbons represent 95% confidence intervals. Shaded ribbons in (J-K) have been cropped to maintain consistent y-axes across the plots, allowing for easier comparison. Sample sizes are indicated (n).

A target of che-1, the guanylyl cyclase gcy-22.3 is required for chemotaxis

We further explored the asymmetric salt perception by the ASE neurons, which in C. elegans is largely mediated through distinct receptor-type guanylyl cyclases (rGC proteins, encoded by gcy genes), which confer salt specificity via their extracellular domains (30) . Our previous genome survey of Ppa homologs of gcy genes has revealed patterns that made us question whether Ppa gcy genes are convincing candidates for lateralized chemotactic responses. Specifically, we noted that ASER-expressed C. elegans gcy genes and ASEL-expressed C. elegans gcy genes have only expanded in the Caenorhabditis genus (11). One outlier to this pattern is the Cel-gcy-22 gene, which is expressed in ASER, but has not expanded in C. elegans (11, 42). However, this gene has duplicated several times in P. pacificus, resulting in 5 putative Ppa-gcy-22 paralogs (11). We fused the promoter of one of these paralogs, Ppa-gcy-22.3, to gfp and found that transgenic animals express GFP exclusively in ASER (Fig. 6A), identical to the C. elegans gcy-22 ortholog (42). We confirmed its expression in ASER by analyzing animals that carry both the Ppa-gcy-22p::gfp reporter and the Ppa-che-1p::rfp reporter, showing a unilateral overlap of these reporters in ASER (Fig. 6B).

The laterally asymmetric expression of gcy-22.3 is dependent on the zinc finger transcription factor CHE-1.

(A) The gcy-22.3::GFP marker is expressed exclusively in the right ASE neuron (ASER)(n>200). (B) The gcy-22.3::GFP marker co-localizes with che-1::RFP expression in the ASER. (C) gcy-22.3::GFP expression is absent in the che-1(ot5012) mutant (n=55). Anterior is left and dorsal is top. Scale bar in (A) represents 50 µm for all panels.

To assess whether Ppa-gcy-22.3 is a potential effector of Ppa-che-1 function, we crossed the gcy-22.3 reporter into che-1(ot5012) mutants. We found expression of gcy-22.3 was eliminated (Fig. 6C), leading us to conclude that gcy-22.3 is a potential effector of che-1 function, identical to its homolog in C. elegans.

To determine whether and which of the observed salt responses is mediated by the ASER-expressing gcy-22.3, we generated a putative gcy-22.3 null mutant through CRISPR/Cas9 genome editing (2 bp complex deletion that introduces a frameshift)(Fig. 7A) and examined its response to the higher salt concentration. The ‘OFF’ response to 250 mM NH4Cl was notably abolished in the ASER neuron in the loss-of-function gcy-22.3 mutant, while the ‘ON’ response in the ASEL remained intact (Fig. 7B-C). However, the responses to 250 mM NaCl were not significantly reduced in the either the ASEL or ASER neuron in the gcy-22.3 mutant (Fig. 7D-E). Furthermore, the gcy-22.3 mutation also reduced the ‘ON’ portion of the AFD biphasic response following presentation of 250 mM NH4Cl and NaCl (Fig. 7F-G). We also examined the behavioral responses toward individual salt ions in gcy-22.3 mutants, including a second loss-of-function allele csu182. Overall, we found no significant differences between wildtype and mutants in responses towards individual ions, and only the attraction to NH4Cl was slightly enhanced in both alleles (Fig. 8), which was unexpected given the lack of calcium response observed in gcy-22.3(csu181) towards NH4Cl. Our findings show that while proper ASE function is critical for salt attraction, defects in individual gcy genes do not lead to a major impact on the worms’ ability to track toward attractive salts.

ASE and AFD responses in wildtype compared to gcy-22.3 mutants.

Average percent change in RCaMP fluorescence (dF/F) over time (seconds) as described in Fig. 4. (A) www.pristionchus.org Genome Browser view of the gcy-22.3 locus with the CRISPR/Cas9-induced mutations indicated with a red bar (Exon 5 of PPA04454 or Exon 4 of Contig12-snapTAU.506). (B-C) ASEL and ASER neuron responses to 250 mM NH4Cl wildtype (gray) and gcy-22.3 mutants (magenta). (D-E) ASEL and ASER neuron responses to 250 mM NaCl in wildtype (gray) and gcy-22.3 mutants (magenta). AFD (combined left and right neurons) responses (F) to 250 mM NH4Cl and (G) to 250 mM NaCl in wildtype (orange) and gcy-22.3 mutants (magenta). Shaded ribbons represent 95% confidence intervals. Sample sizes are indicated (n). For ASEL/R comparisons, bar plots represent the difference between the minimum % dF/F value 10s pre-stimulus and maximum % dF/F value 10s post-stimulus for either the (B) ON or (C-E) OFF stimulus. For AFD comparisons, bar plots represent maximum % dF/F values 10s after the ON and OFF stimulus. *P<0.05, **P<0.01, ***P<0.001 indicate significant difference. “ns” indicate no significant difference. Comparisons between different genotypes were analyzed using unpaired t-test or Mann-Whitney test.

Chemotaxis responses to individual ions in gcy-22.3 mutants.

Young adult hermaphrodite responses to salt gradients with or without various salts in background. *P<0.05 Significant differences were found between wildtype PS312 and gcy-22.3 mutants to NH4Cl by two-way ANOVA and Dunnett’s test. Each assay involves a minimum of 10 animals. Sample sizes for each condition are indicated on the bottom. Error bars denote standard error of the mean.

Discussion

Our study has revealed several insights into the substrates of evolutionary changes between two distantly related nematode species, P. pacificus and C. elegans. We used intracellular calcium levels as our readout for neuronal activity with a genetically encoded calcium sensor in two pairs of che-1-expressing amphid sensory neurons– the first calcium imaging study in P. pacificus. We show that three neuron types (ASE left, ASE right and AFD neurons) each have distinct calcium responses to specific ion concentrations, revealing diversity at the single neuron level. We have identified the first laterally asymmetric marker in P. pacificus, gcy-22.3p::GFP, with its expression limited to ASER homolog (AM7). Our unexpected discovery that neuronal asymmetry is present in the ASE homologs between two distantly related nematode species, despite the lack of a lsy-6 homolog, suggests that functional lateralization in P. pacificus may be mediated by a different genetic pathway compared to C. elegans.

P. pacificus and C. elegans have diverged taste palates

We have shown that while C. elegans is attracted to acetate salts, P. pacificus avoids these acetates. Previous studies have shown that ammonium acetate (NH4Ac) is sensed both as a water-soluble compound as well as a volatile odorant and is mediated by different signaling pathways (36). Like C. elegans, it is therefore likely that ammonium and acetate ions involve a different set of neurons in P. pacificus (i.e. non-che-1 expressing neurons), based on the finding that histamine-treated che-1p::HisCl1 animals did not significantly attenuate their repulsive response to NaAc. We find that sodium chloride is a common attractant for both P. pacificus and C. elegans, although the magnitude of response is lower than previously published results (29). This result is likely due to differences in generating the salt gradients. Nevertheless, we find that P. pacificus neurons have a distinct response to this salt when compared to those observed in C. elegans neurons. The P. pacificus ASEL neuron responds to a decrease in NaCl concentration (likely sodium), as evidenced by the ’OFF-2’ response profile, whereas the C. elegans ASEL neuron responds to an increase in sodium concentration. However, in both nematodes, the ASER neuron responds to a decrease in chloride concentration. Additionally, the P. pacificus ASER neuron exhibits a unique ON-OFF response to ammonium iodide, whereas in C. elegans, no ON-OFF type of response is seen in the ASE neurons– the C. elegans ASER neuron responds only to a decrease in iodide concentration. Based on our calcium imaging and chemotaxis results on the gcy-22.3 mutants, attraction to ammonium ions (NH4+) is likely mediated in part by the ASER neuron in P. pacificus. Pristionchus species are entomophilic and most frequently found to be associated with beetles in a necromenic manner and thus insect cadavers could be sources of ammonium in the soil (6, 43–45) . Additionally, ammonium salts could represent a biological signature of other nematodes that the predatory morph of P. pacificus could interpret as prey. In P. pacificus, nutritional state has a measurable role in the mouth-form polyphenism decision between predatory and non-predatory morphs (46). Collectively, our findings highlight the divergence of the P. pacificus salt-sensory neurons compared to those observed in C. elegans.

P. pacificus ASE neurons have narrower sensitivity range

The sensitivity range of the P. pacificus ASER responses is significantly less compared to the ASEL responses, as well as to C. elegans ASER responses. Whereas the C. elegans ASER has a 40-fold sensitivity range in the ‘OFF’ response to the removal of various concentrations of NaCl (1-40 mM)(29, 47–49), the P. pacificus ASER showed the ’OFF’ response only to 250 mM NH4Cl but not to a 10-fold reduction in concentration of NH4Cl (25 mM). For the ASEL in contrast, the response to 25 mM was just as strong as to 250 mM NH4Cl (10-fold) and comparable to the 8-fold concentration range observed for C. elegans ASEL towards NaCl. Alternatively, the magnitude of these sensitivity differences may be also partially due to differences among calcium indicators (i.e. GCaMP and Cameleon), but multiple P. pacificus che-1p::GCaMP strains did not exhibit sufficient basal fluorescence to allow for image tracking and direct comparison. The narrower sensitivity range in P. pacificus chemosensation was also observed for attraction to volatile odors (2), which span only 10-fold, versus up to 10,000-fold in attractive odors for C. elegans (21).

P. pacificus taste neurons exhibit a unique biphasic response

Although the left-ON and right-OFF responses are conserved in the ASE neurons in both species, the biphasic response of ASER to 250 mM NH4I has not been observed in either of the ASE neurons towards salts. In C. elegans, hyperosmotic stimulus such as 1M glycerol, or high concentrations or long duration of CuSO4 exposure both result in biphasic responses by the ASH neurons that sense noxious chemicals (41, 50). Olfactory neurons that mediate avoidance behavior such as the AWB neurons can also exhibit a biphasic response to the presence and removal of a high concentration isoamyl alcohol that normally elicits avoidance behavior (51). The neuron-specific response can also be dependent on the concentration of the chemical compound, since the ASER to 25 mM NH4I was a weak ‘OFF’ rather than a biphasic one. In contrast to the salt-dependent response types by P. pacificus ASE neurons, the P. pacificus AFD neurons also show exclusively biphasic responses with various amplitudes. Further characterization will help determine if biphasic responses can also be found in other sensory neuron types in P. pacificus, specifically those neurons mediating avoidance behavior.

AFD are potentially polymodal neurons

Broadly speaking, C. elegans chemosensory neurons have been classically characterized as specialized neurons for dedicated modalities such as water-soluble chemicals (ASE), volatile odorants (AWA, AWB, AWC), noxious chemicals (ASH), pheromones (ADL, ADF, ASK), as well as light (ASJ) and temperature (AFD)(1, 37, 52–55). Advances in multi-neuron calcium recordings have since shown that a given odor within a certain concentration range is detected by different ensembles of the 12 amphid neuron classes, including the AFD neurons (56–58). Unexpectedly, we found that the P. pacificus AFD neurons exhibit a distinctive biphasic response to all three salts tested (NH4Cl, NaCl, and NH4I), which differ from the C. elegans AFD calcium responses (58, 59). Moreover, the loss of the receptor gcy-22.3 reduced the AFD ‘ON’ response to NH4Cl and NaCl, indicating ASER contributes to the AFD response. The strong positive AFD response to 25 mM NaCl that is absent in both ASE neurons further supports the likelihood that AFD receives inputs from other amphid neurons. The integration of thermosensation and chemosensation is important for memory-regulated behavior. In C. elegans, maximum chemotaxis indices toward NH4Cl occurs when there is concordance between cultivation temperature and assay temperature (60). The C. elegans AFD neurons are also important for gustatory aversive learning in NaCl avoidance (49). Given the influence of environmental temperature on the P. pacificus mouth-form plasticity and the wide range of micro-climates that wild strains of P. pacificus have been isolated from (61–63), temperature and taste preferences could be regulated at multiple genetic levels during crucial developmental decisions.

Changes in the gene regulatory architecture of sensory neuron specification

We found that the key regulator of Cel ASE identity, che-1, is also expressed in Ppa ASE and may play a similar role as a terminal selector in this neuron type, based on its effect on ASE-mediated behavior and regulation of the Ppa-gcy-22.3 gene. However, the stronger behavioral effect of silencing of che-1 expressing neurons compared to a che-1 mutant background could either indicate that che-1 does not have as broad a role in controlling ASE differentiation in P. pacificus versus C. elegans. It is possible that a developmental loss of ASE differentiation may result in compensatory changes in the chemosensory system during early development, as was observed in the C. elegans mating pheromone response by males (64). It is also worthwhile to note that Cel-gcy-22 stands out for being located on a separate chromosome (Chr. V) compared to the other ASER-type gcy genes (gcy-1, gcy-4, gcy-5 on Chr. II) as well as the Cel-gcy-22 mutant having defects in chemoattraction towards a wide range of salt ions (30, 42). Given that the 5 P. pacificus gcy-22-like paralogs are located on 3 separate chromosomes (Chr. I, IV, X), it is likely that they emerged from independent and repeated gene duplication events after the separation of Caenorhabditis and Pristionchus lineages. Although the promoter fusion reporters of other gcy genes have been uninformative due to lack of expression (Ppa-gcy-22.1, Ppa-gcy-7.1, Ppa-gcy-7.2, Ppa-gcy-5), it is likely that ASEL-specific gcy genes as well as additional ASER-specific gcy-22-like genes exist. Finding other genes with left-right specific expression could help to identify genetic determinants affecting lateral asymmetry.

Unexpectedly, we found that unlike in C. elegans, the Ppa-che-1 gene is also expressed in the AFD neurons. Since we cannot record neural activity in AFD in a che-1 mutant (the che-1p::RCaMP driver fails to be expressed sufficiently in che-1 mutants due to autoregulation), and do not yet have molecular markers for Ppa AFD neurons, we cannot assess whether che-1 affects AFD neuron differentiation.

Perhaps the most striking difference in the gene regulatory architecture of ASE neuron specification is the apparent lack of the key regulator of ASE asymmetry in P. pacificus, the miRNA lsy-6. In C. elegans, the expression of lsy-6 exclusively in ASEL is prepatterned via an early embryonic Notch signal (33) and serves to downregulate the homeodomain transcription factor cog-1 in the ASEL neuron (65). Through a network of downstream regulatory events, asymmetry of rGCs eventually becomes established (31). cog-1 and several asymmetrically expressed downstream effectors of cog-1, such as the die-1, lim-6 and fozi-1 transcription factors are conserved in P. pacificus, but whether the function of these factors in controlling P. pacificus ASE laterality is conserved remains to be determined. In this context it is intriguing to note that another prominent gene regulatory pathway that is controlled by miRNAs in C. elegans, the heterochronic pathway (let-7s, lin-4), appears to have diverged in P. pacificus as well, despite the conservation of the overall physiological readouts of this pathway (temporal patterning of cell lineage divisions)(66). It is tempting to speculate that miRNA-meditated regulatory process is particularly labile.

In conclusion, our work illustrates how comparative behavioral and genetic analyses in nematodes is a powerful strategy to uncover substrates of evolutionary change in simple nervous systems.

Materials and Methods

Nematode Strains

P. pacificus and other nematode strains were maintained at ∼20°C on NGM plates seeded with E. coli OP50 for food as described previously (67); these are derived from standard C. elegans culture methods (68). P. pacificus and other nematode strains used are listed in Table S1.

Chemotaxis assays

The assay for assessing response to salt gradients was adapted from C. elegans and P. pacificus chemotaxis assays (2, 21, 42). Overnight salt gradients were established on 10 cm chemotaxis plates containing 20 ml agar (5 mM KPO4, 1 M CaCl2, 3% Bacto-agar, 1 mM MgSO4) by adding 10 µl of 2.5 M salt solutions for 16 hours. Alternatively, agar containing 25 mM (NH4Cl, NH4I) or 50 mM (NaCl) were used to test for responses to individual salt ions. Following the establishment of the overnight point gradient, another 4 µl of the same salt solution or water control was added to reinforce the gradient 4 hours before the assay. Just prior to the assay, 1 µl of 1 M sodium azide was added to both the attractive salt (A) and the control (C) spots. P. pacificus J4 to adult hermaphrodites from near-saturated cultures were washed 3x with distilled water and collected by centrifuging at 2000 rpm for 2 minutes. Approximately 200 worms were loaded onto the edge of each assay plate between the gradient sources, and at least 10 combined worms have to reach the scoring arenas to be considered. At least 10 assays constituted each experimental trial, and multiple trials were conducted and averaged for each condition. The Chemotaxis Index (CI) for each end-point assay plate is defined as (A - C)/(A + C). To conduct conditional knockdowns of neurons, 5 M histamine dihydrochloride (Sigma Aldrich H7250) stock solution in sterilized deionized water (Arrowhead CA) was filter-sterilized and top plated onto the agar plates to a final histamine concentration of 25 mM histamine approximately 10 minutes before loading the worms to commence the assay. Most assays lasted 3-4 hours at room temperature to allow P. pacificus sufficient time to reach the scoring arenas, with ∼40% of the animals participating. When using the csuEx83[Ppa-che-1p::optHisCl1] strain, only animals expressing Ppa-egl-20p::RFP tail marker from the extrachromosomal array were scored. Because of P. pacificus’ strong aversion to acetate, we could not easily assess the individual contributions of salt ions in a saturated background of ammonium acetate as conventionally practiced in C. elegans studies (30).

Ppa-che-1p::HisCl1 strain

To make the Ppa-che-1p::optHisCl1, the P. pacificus codon-optimized histamine-gated chloride channel sequence used in C. elegans (HisCl1)(69, 70) was designed using (https://hallemlab.shinyapps.io/Wild_Worm_Codon_Adapter/)(71) and custom synthesized (Twist Bioscience), and subsequently inserted behind the che-1 promoter (3.1 kb containing the first exon and intron)(11) to create the pHC30 plasmid construct. This Ppa-che-1p::optHisCl1 plasmid (2 ng/µl) along with PS312 genomic DNA (80 ng/µl) and Ppa-egl-20p::RFP co-injection marker (2 ng/µl) were individually digested with HindIII and assembled as the injection mix to create csuEx83.

Ppa-che-1p::RCaMP reporter strain

To make the Ppa-che-1pei::optRCaMP, we generated a transgenic worm strain expressing the codon-optimized genetically-encoded calcium indicator (GECI), jRCaMP1a, in the neurons of interest (72, 73). jRCaMP1a is an improved red GECI based on mRuby with comparable sensitivity to GCaMP6 (74). The codon-optimized RCaMP sequence was custom synthesized (Twist Bioscience), and then subcloned using the plasmid pJET1.2/blunt (Thermo Fisher Scientific) to create the pMM2 plasmid construct. Using Gibson Assembly (E2611, New England Biolab), the codon-optimized RCaMP sequence was introduced downstream of the Ppa-che-1 promoter sequence with the first exon and intron sequences to create the pMM5 plasmid construct. Separately, we also generated a Ppa-che-1pei::optGFP transcriptional reporter with codon-optimized GFP (pMM3) to enable the localization of the che-1-expressing neurons during video acquisition (75). The Ppa-che-1pei::optRCaMP (2 ng/µl) and Ppa-che-1pei::optGFP constructs (1 ng/µl), along with PS312 genomic DNA (80 ng/µl) and Ppa-egl-20p::RFP (1.5 ng/µl) were individually digested with HindIII and assembled as the injection mix to create csuEx93. Despite multiple attempts, we were unable to generate an equivalent che-1pei::GCaMP transgenic line with sufficient basal level of GCaMP expression for a comparison to the RCaMP calcium dynamics.

Promoter fusion reporter strains

The Ppa-gcy-22.3p::GFP construct contains 899 bp of the upstream promoter of the PPA04554 transcript is fused to the codon-optimized GFP (pVL2). A shorter gcy-22.3 transcript with different first two exons is also predicted, which is more similar in length to other rGC genes in P. pacificus (Contig12-snapTAU.506). The Ppa-ttx-1pei::RFP construct contains 1935 bp of the upstream promoter along with the first exon and intron sequences of Ppa-ttx-1 (PPA26714) fused to the codon-optimized RFP, excluding the first 2 codons (pDC14). pVL2 (2 ng/µl) or pDC14 (2 ng/µl), along with PS312 genomic DNA (80 ng/µl) were individually digested with HindIII to create the injection mixes to generate the independent reporter strains, csuEx90[Ppa-gcy-22.3p::GFP], as well as Ppa-ttx-1pei::RFP(csuEx94) and Ppa-ttx-1pei::RFP(csuEx96). csuEx96 showed less gland cell expression (which can occlude neuronal expression) and stronger AFD expression than csuEx94.

CRISPR mutagenesis generated mutants

CRISPR/Cas9 mutagenesis was used to generate mutations (75, 76). crRNA and primer sequences, and induced mutations, are included in Table S3 and Fig. S8.

che-1 alleles (PPA01143)

Target crRNA, tracrRNA, and Cas9 nuclease were purchased from IDT Technologies (San Diego, CA). crRNA and tracrRNA were hydrated to 100 µM with IDT Duplex Buffer, and equal volumes of each (0.61 µl) were combined and incubated at 95°C for 5 minutes, then 25°C for 5 minutes. Cas9 protein (0.5 µl of 10 µg/µl) was added, then the mix was incubated at 37°C for 10 minutes. Ppa-egl-20p::RFP was used as a co-injection marker. To reach a final total volume of 40 µl, the Cas9-crRNA-tracrRNA complex was combined with pZH009 (Ppa-egl-20p::RFP) DNA to reach 50 ng/µl final concentration using nuclease-free water. F1 progeny were screened for the presence of Ppa-egl-20p::RFP expression in the tail and candidate F1‘s were sequenced to identify heterozygotes (76). ot5012 has a 4 bp insertion while ot5013 has an 8 bp complex insertion/deletion, and both mutations cause frameshift mutations and premature stop codons. Each allele was outcrossed two times to wildtype before characterization.

gcy-22.3 alleles (PPA04454)

Target crRNA, tracrRNA, and Cas9 nuclease were purchased from IDT Technologies (San Diego, CA). crRNA (RHL1400) and tracrRNA were hydrated to 100 µM with IDT Duplex Buffer, and equal volumes of each (0.61 µl) were combined and incubated at 95°C for 5 minutes, then 25°C for 5 minutes. Cas9 protein (0.5 µl of 10 µg/ µl) was added, then the mix was incubated at 37°C for 10 minutes. Ppa-egl-20p::RFP was used as a co-injection marker. To reach a final total volume of 40 µl, the Cas9-crRNA-tracrRNA complex was combined with pZH009 (Ppa-egl-20p::RFP) DNA to reach 50 ng/µl final concentration using nuclease-free water. F1 progeny were screened for the presence of Ppa-egl-20p::RFP expression in the tail and candidate F1‘s were sequenced to identify heterozygotes (76). csu181 has a 2 bp complex deletion, while csu182 has a 22 bp complex deletion, both mutations cause frameshifts and premature stop codons. Each allele was outcrossed two times to wildtype before characterization.

ALFA C-terminal tagging and immunostaining

For induction of site-specific insertions via CRISPR/Cas-9 mediated mutagenesis, target crRNA, tracrRNA, and Cas9 nuclease were treated as described above. Single-stranded DNA repair template containing the ALFA nanobody tag (RHL1551 for CHE-1 and RHL1519 for TTX-1) with 35 bp of homology arms on the 5’ and 3’ sides were purchased from IDT. To minimize sequence identity, the two copies of the ALFA sequence contain silent mutations. The crRNA (RHL1396 for che-1 and RHL1514 for ttx-1) and tracrRNA were hydrated to 100 µM with IDT Duplex Buffer, and equal volumes of each (0.61 µl) were combined and incubated at 95°C for 5 minutes, then 25°C for 5 minutes. The ALFA tag was C-terminally inserted to the longest ttx-1 splice form in the last 18th exon (ppa_stranded_DN30925_c0_g1_i5)(Trinity 2016 transcripts can be found on www.pristionchus.org). F1 animals expressing the co-injection marker egl-20::optRFP were lysed and checked by PCR for insertions.

Non-starved healthy cultures of ALFA-tagged CHE-1 (csu226[Ppa-che-1::2xALFA]) or ALFA-tagged TTX-1 (RLH280[Ppa-ttx-1::2xALFA]) (from six 6 cm plate cultures) were washed with M9 and filtered (Sartorius 84g/m2, Grade 392) and processed as previously described (39). In brief, mixed stage worms were fixed overnight at 4°C on a nutator with 500 µl fixation buffer (4% paraformaldehyde in PBS). The worms were then incubated overnight at 37°C with 500 µl 4% β-mercaptoethanol dissolved in 1% Triton-X100 in 0.1M Tris pH 7.4, and digested in 200 μl of collagenase buffer with 200 units of collagenase type IV (Gibco, Thermo Fisher Scientific, Waltham, MA, USA) at 37°C for ∼3.5 hours. The partially digested worms were subsequently washed three times with 500 µl PBST. After collagenase treatment the centrifugation between the washes was done at low speed, 1000 RCF. After washes, the worms were stained in 50 µl 1% BSA in PBST with the primary antibody (1:100 FluoTag-X2 anti-ALFA-AZdye568, N1502, NanoTag Biotechnologies, Göttigen, Germany), overnight at 4°C in a nutator. Following the washes, worms were resuspended in 50 µl VectaShield mounting medium (H-1000, Vector Laboratories, USA) containing DAPI (Molecular Probes, Thermo Fisher Scientific) and gently mounted on freshly prepared 3% Noble agar pads. Images were acquired on a Leica DM6000 microscope.

Calcium Imaging

To conduct calcium imaging, worms were trapped and imaged within a microfluidic PDMS chip while delivering stimuli directly to the nose of an immobilized worm, as previously described (50, 77). The microfluidic chip was connected to a programmable valve controller (ValveBank) that enables the user to toggle between “stimulant ON” and “stimulant OFF” states. Specifically, the ValveBank allows controlled switching of flow from the two outer buffer channels in the chip such that either the control solution (stimulant OFF) or the stimulant solution (stimulant ON) flows over the worm nose (Fig. S9). Buffer solution (for the outer two channels) consisted of M9 buffer with 0.1% Tween-20 and 1 µg/ml fluorescein (78, 79). The worm loading solution consisted of M9 buffer with 0.1% Tween-20 and 1.5 mM tetramisole hydrochloride to immobilize the animal. Water-soluble stimulant solutions consisted of the following water-soluble compounds dissolved in nanopure/milli-Q water: ammonium chloride (NH4Cl), ammonium iodide (NH4I), and sodium chloride (NaCl) at concentrations of 250 mM and 25 mM. 750 mM NaCl elicited very inconsistent responses. The water-soluble stimulant solution was made by diluting the water-soluble compound in nanopure/milli-Q water with 0.1% Tween-20. The control solution consisted of 0.1% Tween-20 nanopure/milli-Q water. Worms were exposed to the control solution, stimulant solution, and then control solution using a 60-second program: 10 seconds stimulant OFF, 20 seconds stimulant ON, and 30 seconds stimulant OFF. As a negative control, animals were exposed only to the control solution, without switching channels, for the duration of the recording. For each animal, the orientation of the nose and vulva were recorded and used as a guide to determine the ventral and dorsal sides of the worm, and subsequently, the left and right sides of the worm. Accounting for the plane of focus of the neuron pairs as viewed through the microscope, it was then determined whether the imaged neuron was the worm’s left or right neuron of each pair. Images were captured using a Zeiss Axio Observer Z1 inverted fluorescence microscope and a pco.panda 4.2 sCMOS camera. Changes in fluorescence intensity were measured in the neurons of interest while the worm was exposed to green light. Images were processed using MetaMorph software version 7.10.5.476. Images were captured at 500 ms exposure time at 2 fps because the baseline fluorescence in Ppa-che-1p::optRCaMP worms was too dim to capture viable data using the standard 100 ms exposure time. Baseline F₀ was measured as the average background-subtracted fluorescence from the first 9 seconds of each recording and change in fluorescence intensity was calculated as dF/F = (F - F₀) / F₀, as described (79). Data were analyzed and plotted using custom scripts generated in MATLAB versions R2021a–R2024a. This code is available at https://github.com/honglabcsun/Calcium-Imaging.

For bar plot comparisons between wildtype and gcy-22.3 mutants, we calculated minimum pre-stimulus and maximum post-stimulus % dF/F values using custom scripts generated in MATLAB. Min-Max data were exported as text files, manually converted and organized into an XLSX file (Microsoft Excel Office16) and imported into Prism GraphPad software (version 10) for data visualization and statistical analysis.

Nomenclature

Throughout the results section, P. pacificus genes will be referred to without the Ppa- prefix; if necessary for comparison to another species such as C. elegans (Cel-) the Ppa- prefix will then be used.